---
title: Livestock Breeding and Genomics - Solution 9
author: "`r Sys.info()[['user']]`"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Problem 1: Decompositions
Given is the following pedigree. 

```{r pedsol09p01, echo=FALSE, results='asis'}
n_nr_pedsol09p01 <- 6
tbl_pedsol09p01 <- tibble::data_frame(Animal = 1:n_nr_pedsol09p01,
                                      Sire   = c(NA, NA, 1, 1, 3, 5),
                                      Dam    = c(NA, NA, 2, NA, 4, 4))
knitr::kable(tbl_pedsol09p01,
             booktabs = TRUE, 
             longtable = TRUE)
```

### Your Tasks
* Do the simple decomposition of the above pedigree
* Do the recursive decomposition of the above pedigree until only $m_i$ terms appear on the right-hand side of the decomposition.
 
### Solution

* __Simple Decomposition__: For the simple decomposition, the true breeding values are decomposed into true breeding values of parents plus the respective mendelian sampling effect. For the pedigree given above this is

\begin{align}
a_1  &=  m_1  \notag \\
a_2  &=  m_2  \notag \\
a_3  &=  {1\over 2}a_1 + {1\over 2}a_2 + m_3 \notag \\
a_4  &=  {1\over 2}a_1 + m_4 \notag \\
a_5  &=  {1\over 2}a_3 + {1\over 2}a_4 + m_5 \notag \\
a_6  &=  {1\over 2}a_5 + {1\over 2}a_4 + m_6 \notag
\end{align}

Converting the same decomposition into matrix-vector notation, we get

$$a = P \cdot a + m$$

Putting the information from the pedigree into the decomposition yields

```{r simpledecomppedsol09p01, echo=FALSE, results='asis'}
vec_a_sol09p01 <- rmddochelper::vecGetVecElem(psBaseElement = "a", pnVecLen = n_nr_pedsol09p01, psResult = "latex")
vec_m_sol09p01 <- rmddochelper::vecGetVecElem(psBaseElement = "m", pnVecLen = n_nr_pedsol09p01, psResult = "latex")
mat_p_sol09p01 <- matrix(c(rep(0, n_nr_pedsol09p01),
                           rep(0, n_nr_pedsol09p01),
                           0.5, 0.5, rep(0, (n_nr_pedsol09p01-2)),
                           0.5, rep(0, (n_nr_pedsol09p01-1)),
                           0, 0, 0.5, 0.5, 0, 0,
                           0, 0, 0, 0.5, 0.5, 0), nrow = n_nr_pedsol09p01, byrow = TRUE)
### # show matrix vector notation
cat("$$\n")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" = ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_p_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" \\cdot ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" + ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_m_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


* __Recursive Decomposition__: The recursive decomposition repeats simple decompositions of true breeding values of ancestors until the right-hand side of the decomposition consists only of mendelian sampling terms. In matrix-vector notation, the recursive decomposition can be written as

$$a = L \cdot m$$

The vectors $a$ and $m$ are defined as for the simple decomposition. The matrix $L$ has the following structure.




## Problem 2: 


### Solution

