---
title: Livestock Breeding and Genomics - Solution 9
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Problem 1: Decompositions
Given is the following pedigree. 

```{r pedsol09p01, echo=FALSE, results='asis'}
n_nr_pedsol09p01 <- 6
tbl_pedsol09p01 <- tibble::data_frame(Animal = 1:n_nr_pedsol09p01,
                                      Sire   = c(NA, NA, 1, 1, 3, 5),
                                      Dam    = c(NA, NA, 2, NA, 4, 4))
knitr::kable(tbl_pedsol09p01,
             booktabs = TRUE, 
             longtable = TRUE)
```

### Your Tasks
* Do the simple decomposition of the above pedigree
* Do the recursive decomposition of the above pedigree until only $m_i$ terms appear on the right-hand side of the decomposition.
 
### Solution

* __Simple Decomposition__: For the simple decomposition, the true breeding values are decomposed into true breeding values of parents plus the respective mendelian sampling effect. For the pedigree given above this is

\begin{align}
a_1  &=  m_1  \notag \\
a_2  &=  m_2  \notag \\
a_3  &=  {1\over 2}a_1 + {1\over 2}a_2 + m_3 \notag \\
a_4  &=  {1\over 2}a_1 + m_4 \notag \\
a_5  &=  {1\over 2}a_3 + {1\over 2}a_4 + m_5 \notag \\
a_6  &=  {1\over 2}a_5 + {1\over 2}a_4 + m_6 \notag
\end{align}

Converting the same decomposition into matrix-vector notation, we get

$$a = P \cdot a + m$$

Putting the information from the pedigree into the decomposition yields

```{r simpledecomppedsol09p01, echo=FALSE, results='asis'}
vec_a_sol09p01 <- rmddochelper::vecGetVecElem(psBaseElement = "a", pnVecLen = n_nr_pedsol09p01, psResult = "latex")
vec_m_sol09p01 <- rmddochelper::vecGetVecElem(psBaseElement = "m", pnVecLen = n_nr_pedsol09p01, psResult = "latex")
mat_p_sol09p01 <- matrix(c(rep(0, n_nr_pedsol09p01),
                           rep(0, n_nr_pedsol09p01),
                           0.5, 0.5, rep(0, (n_nr_pedsol09p01-2)),
                           0.5, rep(0, (n_nr_pedsol09p01-1)),
                           0, 0, 0.5, 0.5, 0, 0,
                           0, 0, 0, 0.5, 0.5, 0), nrow = n_nr_pedsol09p01, byrow = TRUE)
### # show matrix vector notation
cat("$$\n")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" = ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_p_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" \\cdot ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" + ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_m_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


* __Recursive Decomposition__: The recursive decomposition repeats simple decompositions of true breeding values of ancestors until the right-hand side of the decomposition consists only of mendelian sampling terms. In matrix-vector notation, the recursive decomposition can be written as

$$a = L \cdot m$$

The vectors $a$ and $m$ are defined as for the simple decomposition. The matrix $L$ has the following structure.

```{r recdecomppedsol09p01, echo=FALSE, results='asis'}
suppressPackageStartupMessages( library(pedigreemm) )
ped_sol09p01 <- pedigree(sire = tbl_pedsol09p01$Sire, dam = tbl_pedsol09p01$Dam, label = as.character(tbl_pedsol09p01$Animal))
mat_a_sol09p01 <- as.matrix(getA(ped = ped_sol09p01))
mat_r_sol09p01 <- t(chol(mat_a_sol09p01))
### # verify
mat_d_sol09p01 <- diag(Dmat(ped = ped_sol09p01), nrow = n_nr_pedsol09p01)
mat_s_sol09p01 <-  sqrt(mat_d_sol09p01)
mat_l_sol09p01 <- mat_r_sol09p01 %*% solve(mat_s_sol09p01)
ssq_dev <- sum(sum((mat_a_sol09p01 - mat_l_sol09p01 %*% mat_d_sol09p01 %*% t(mat_l_sol09p01))^2))
if ( ssq_dev > .Machine$double.eps )
  stop("Verification of LDL-Decomposition of A failed with a summed square deviation of: ", ssq_dev, "\n")

### # show the decomposition equation
cat("$$\n")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" = ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_l_sol09p01, pnDigits = 3), collapse = "\n"), "\n")
cat("\\right]\n")
cat(" \\cdot ")
cat("\\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_m_sol09p01), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```



## Problem 2: Covariance Between Animals
So far we have assumed that the covariance between true breeding values $a_i$ and $a_j$ of animals $i$ and $j$ is defined as 

$$cov(a_i, a_j) = (A)_{ij} * \sigma_a^2$$

\begin{tabular}{lll}
where  &  $(A)_{ij}$    &  element of numerator relationship matrix $A$ on row $i$ and column $j$  \\
       &  $\sigma_a^2$  &  additive genetic variance
\end{tabular}


### Your Task
* Compute the covariance $cov(a_i, a_j)$ between true breeding values $a_i$ and $a_j$ for all pairs of animals in the pedigree given in Problem 1.
* The following procedure can be used to compute the covariances
    + Recursively decompose $a_i$ and $a_j$ until common ancestors between $i$ and $j$ are found or until unrelated founder animals occur in the decomposition
    + Expand the covariance between sums of true breeding values into sums of variances of true breeding values
    + Collect the results of all pairwise covariances and compare them to the entries in $A$
    
    
### Hints
* The covariance $cov(a_i, a_j)$ between true breeding values $a_i$ and $a_j$ for unrelated animals $i$ and $j$ is set to $0$. An example in the pedigree from Problem 1 would be the covarance $cov(a_1, a_2)$ between the true breeding values of animals $1$ and $2$ which is equal to $0$, because animals $1$ and $2$ are unrelated, that means they do not have any known common ancestors. 
* The covariances $cov(a_i, m_k)$ between any true breeding value $a_i$ and any mendelian sampling constant $m_k$ is $0$. 
* Any previously compute covariance can be re-used.
* The variances $var(a_i)$ of the true breeding values are $var(a_i) = (1 + F_i) \sigma_a^2$ where $F_i$ is the inbreeding coefficient of animal i.
* The inbreeding coefficients for all animals in pedigree of Problem 1 are

```{r, echo = FALSE, results='asis'}
matA <- as.matrix(getA(ped = ped_sol09p01))
tbl_inbreeding_sol09prob02 <- tibble::data_frame(Animal = 1:n_nr_pedsol09p01,
                                                 `Inbreeding Coefficient` = inbreeding(ped = ped_sol09p01))
knitr::kable(tbl_inbreeding_sol09prob02, 
             booktabs = TRUE,
             longtable = TRUE)
```

* As an example, it is shown how to compute $cov(a_3, a_4)$. First replace $a_3$ and $a_4$ by their simple decomposition into true breeding values of parents plus mendelian sampling terms. Then expand the covariance of two sums into sums of pairwise covariances. Then known values are inserted and the result is computed.

\begin{align}
cov(a_3, a_4)  &=  cov(({1\over 2}a_1 + {1\over 2}a_2 + m_3), ({1\over 2}a_1 + m_4)) \notag \\
               &=  cov({1\over 2}a_1, {1\over 2}a_1) + cov({1\over 2} a_2,  {1\over 2}a_1) \notag \\
               &=  {1\over 4} var(a_1) = {1\over 4} \sigma_a^2 \notag
\end{align}


### Solution

\begin{align}
cov(a_1, a_2) &= 0 \notag
\end{align}

\begin{align}
cov(a_1, a_3)  &=  cov(a_1, {1\over 2}a_1 + {1\over 2}a_2 + m_3) \notag \\
               &=  cov(a_1, {1\over 2}a_1) + cov(a_1, {1\over 2}a_2)  \notag \\
               &=  {1\over 2} cov(a_1, a_1) =  {1\over 2} var(a_1) = {1\over 2} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_1, a_4)  &=  cov(a_1, {1\over 2}a_1  + m_4) \notag \\
               &=  cov(a_1, {1\over 2}a_1)  \notag \\
               &=  {1\over 2} cov(a_1, a_1) =  {1\over 2} var(a_1) = {1\over 2} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_1, a_5)  &=  cov(a_1, {1\over 2}a_3 + {1\over 2}a_4  + m_5) \notag \\ 
               &=  cov(a_1, {1\over 2}({1\over 2}a_1 + {1\over 2}a_2 + m_3) + {1\over 2}({1\over 2}a_1 + m_4) + m_5) \notag \\ 
               &=  cov(a_1, {1\over 4}a_1) + cov(a_1, {1\over 4}a_2  + cov(a_1, {1\over 4}a_1) \notag \\ 
               &=  {1\over 4}cov(a_1, a_1) + {1\over 4}cov(a_1, a_1) = {1\over 2} var(a_1) = {1\over 2} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_1, a_6)  &=  cov(a_1, ({1\over 2}a_5 + {1\over 2}a_4  + m_6)) \notag \\ 
               &=  cov(a_1, ({1\over 2}({1\over 2}a_3 + {1\over 2}a_4  + m_5) + {1\over 2} ({1\over 2}a_1 + m_4) + m_6)) \notag \\
               &=  cov(a_1, ({1\over 2}({1\over 2}({1\over 2}a_1 + {1\over 2}a_2 + m_3) + {1\over 2}({1\over 2}a_1 + m_4)  + m_5) + {1\over 2} ({1\over 2}a_1 + m_4) + m_6 ))\notag \\
               &=  cov(a_1, {1\over 8}a_1) + cov(a_1, {1\over 8}a_2)  + cov(a_1, {1\over 8}a_1) + cov(a_1, {1\over 4}a_1) \notag \\
               &=  {1\over 8} var(a_1) + {1\over 8} var(a_1) + {1\over 4}var(a_1) = {1\over 2} \sigma_a^2 \notag
\end{align}


\begin{align}
cov(a_2, a_3)  &=  cov(a_2, {1\over 2}(a_1 + a_2 + m_3)) \notag \\
               &=  cov(a_2, {1\over 2}a_1) + cov(a_2, {1\over 2}a_2) = {1\over 2} var(a_2) = {1\over 2} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_2, a_4)  &=  cov(a_2, ({1\over 2}a_1 + m_4) \notag\\
               &=  cov(a_2, {1\over 2}a_1) = 0 \notag
\end{align}

\begin{align}
cov(a_2, a_5)  &=  cov(a_2, ({1\over 2}a_3 + {1\over 2}a_4  + m_5)) \notag \\
               &=  cov(a_2, ({1\over 2}({1\over 2}a_1 + {1\over 2}a_2 + m_3) + {1\over 2}({1\over 2}a_1 + m_4) + m_5)) \notag \\
               &=  cov(a_2, {1\over 4}a_1) + cov(a_2, {1\over 4}a_2) + cov(a_2, {1\over 4}a_1)\notag \\
               &=  {1\over 4} var(a_2) = {1\over 4} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_2, a_6)  &=  cov(a_2, ({1\over 2}a_5 + {1\over 2}a_4  + m_6))) \notag \\
               &=  cov(a_2, ({1\over 2}({1\over 2}a_3 + {1\over 2}a_4  + m_5) +  {1\over 2} ({1\over 2}a_1 + m_4) + m_6)) \notag \\
               &=  cov(a_2, {1\over 4} a_3) + cov(a_2, {1\over 4} a_4) + cov(a_2, {1\over 4}a_1)\notag \\
               &=  {1\over 4} cov(a_2, a_3) = {1\over 8} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_3, a_4)  &=  cov(({1\over 2}a_1 + {1\over 2}a_2 + m_3), ({1\over 2}a_1 + m_4)) \notag \\
               &=  cov({1\over 2}a_1, {1\over 2}a_1) + cov({1\over 2} a_2,  {1\over 2}a_1) \notag \\
               &=  {1\over 4} var(a_1) = {1\over 4} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_3, a_5)  &=  cov(a_3, ({1\over 2}a_3 + {1\over 2}a_4  + m_5))) \notag \\
               &=  cov(a_3, {1\over 2}a_3) + cov(a_3, {1\over 2}a_4) \notag \\
               &=  {1\over 2} var(a_3) + {1\over 2} cov(a_3, a_4) =  {1\over 2} \sigma_a^2 + {1\over 8} \sigma_a^2 =  {5\over 8} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_3, a_6)  &=  cov(a_3, ({1\over 2}a_5 + {1\over 2}a_4  + m_6)) \notag \\
               &=  {1\over 2} cov(a_3, a_5) + {1\over 2} cov(a_3, a_4) \notag \\
               &=  {5\over 16} \sigma_a^2 + {1\over 8} \sigma_a^2 = {7\over 16} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_4, a_5)  &=  cov(a_4, ({1\over 2}a_3 + {1\over 2}a_4  + m_5)) \notag \\
               &=  cov(a_4, {1\over 2}a_3) + cov(a_4, {1\over 2}a_4) \notag \\
               &=  {1\over 2} cov(a_4, a_3) + {1\over 2} var(a_4) = {1\over 2}*{1\over 4} \sigma_a^2 + {1\over 2} \sigma_a^2 =  {5\over 8} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_4, a_6)  &=  cov(a_4,  ({1\over 2}a_5 + {1\over 2}a_4  + m_6))) \notag \\
               &=  cov(a_4, {1\over 2}a_5) + cov(a_4,  {1\over 2}a_4) \notag \\
               &=  {1\over 2}cov(a_4, a_5) + {1\over 2} var(a_4) = {1\over 2} * {5\over 8} \sigma_a^2 + {1\over 2} \sigma_a^2 = {13 \over 16} \sigma_a^2 \notag
\end{align}

\begin{align}
cov(a_5, a_6)  &=  cov(a_5, ({1\over 2}a_5 + {1\over 2}a_4  + m_6))) \notag \\
               &=  {1\over 2} cov(a_5, a_5) +  {1\over 2}cov(a_5, a_4) \notag \\
               &=  {1\over 2} var(a_5) + {1\over 2}*{5\over 8} \sigma_a^2 = {1\over 2} * (1 + {1\over 8}) \sigma_a^2 + {5\over 16} \sigma_a^2 = {7\over 8} \sigma_a^2 \notag
\end{align}

The results of all the covariances is verified by the fact that 

$$var(a) = A * \sigma_a^2$$

and the factors besides $\sigma_a^2$ are the same as found in the numerator relationship matrix $A$. 

```{r, echo = FALSE, results='asis'}
mat_a_sol09p02 <- as.matrix(getA(ped = ped_sol09p01))
### # show the matrix
cat("$$\n")
cat("A = \\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_a_sol09p02, pnDigits = 4), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

