---
title: "Special Matrices"
author: "Peter von Rohr"
date: "10/19/2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Transpose

* For any matrix $A_{m,n}$ its __transpose__ is called $A^T$
* $A^T$ is a matrix with $n$ rows and $m$ columns
* Element $(A)_{ij}$ in $A$ corresponds to element $(A^T)_{ji}$ in $A^T$
* In R:

```{r transposer, echo=TRUE}
(A <- matrix(c(1:6), nrow = 3, ncol = 2, byrow = TRUE))
t(A)
```


## Symmetric Matrices

* Matrices which are identical to its transpose are called __symmetric__
* Hence for a symmetric matrix $S$

$$S = S^T$$

```{r computesymmatrix}
S <- crossprod(A)
```

```{r showsymmmatrix, echo=TRUE}
S
t(S)
```




## Identity Matrix

* Name: $I_n$
* Dimensions: $n\times n$
* Example $n=3$

```{r identiylatex, results='asis'}
cat("$$I_3 = \\left[ ")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = diag(1,nrow = 3, ncol = 3)), sep = "\n"), "\n")
cat("\\right] $$\n")
```

* In R:

```{r identityr, echo=TRUE}
diag(1, nrow = 3, ncol = 3)
```


## Importance of $I_n$

* Neutral element of matrix multiplication
* For any matrix $A_{m,n}$:

$$ A \cdot I_n = I_m \cdot A = A$$


## Diagonal Matrices

* Matrices with non-zero elements only on diagonal
* Diagonal matrices are symmetric
* In R:

```{r diagonalmatrixr, echo=TRUE}
D <- diag(c(5,8,-3))
D
```


## Inverse Matrix

* The matrix $B$ is called the __inverse__ of matrix $A$, if their product $A \cdot B$ is the identity matrix $I$

$$A \cdot B = I$$
* If it exists, the inverse matrix of matrix $A$ is called $A^{-1}$
* In R the inverse of matrix $S$ is computed as:

```{r inversematr, echo=TRUE}
solve(S)
```



