---
title: "Understanding Variance Analysis Tables"
output: html_notebook
---


## Disclaimer
In Mrode (chpt 11), there is a variance analysis given by the following table.


## Anova

\begin{tabular}{lll}
\hline \\
Source           &  Degrees of Freedom ($df$)          &  Sums of Squares ($SSQ$) \\
\hline \\
Overall ($\mu$)  &  $Rank(X)=1$                  &  $y^TX(X^TX)^{-1}X^Ty = F$  \\
Sires ($s$)      &  $Rank(Z) - Rank(X) = q - 1$  &  $y^TZ(Z^TZ)^{-1}Z^Ty - y^TX(X^TX)^{-1}X^Ty = S$  \\
Residual ($e$)   &  $n - Rank(Z) = n - q$        &  $y^Ty - y^TZ(Z^TZ)^{-1}Z^Ty = R$ \\
\hline \\
Total            &  $n$                          &  $y^Ty$ \\
\hline
\end{tabular}


## Numerical Example

```{r}
tbl_num_ex_chp12 <- tibble::data_frame( Animal = c(4, 5, 6, 7),
                                        Sire   = c(2, 1, 3, 2),
                                        WWG    =  c(2.9, 4.0, 3.5, 3.5) )
knitr::kable(tbl_num_ex_chp12,
             booktable = TRUE, 
             longtable = TRUE)

```

The design matrices and the observation vector corresponds to

```{r}
n_nr_obs_p02 <- nrow(tbl_num_ex_chp12)
### # design matrix X
mat_x_p02 <- matrix(1, nrow = n_nr_obs_p02, ncol = 1)
### # design matrix Z
mat_z_p02 <- matrix(c(0, 1, 0,
                      1, 0, 0,
                      0, 0, 1,
                      0, 1, 0), nrow = n_nr_obs_p02, byrow = TRUE)
n_nr_sire <- ncol(mat_z_p02)
### # Observations
mat_obs <- matrix(tbl_num_ex_chp12$WWG, ncol = 1)

(ytx <- crossprod(mat_obs,mat_x_p02))
ytx == sum(mat_obs[,1])
```

Hence from the above, we can see that $y^TX$ is just the sum of all observations which is contained in the formula for $F$.

Looking at the formula for $S$, we get

```{r}
(ytz <- crossprod(mat_obs, mat_z_p02))
tcrossprod(ytz)
```

which means $y^TZZ^Ty$ corresponds to the sum of the squared observations within each sire, summed over all sires. The multiplication by $(Z^TZ) divides the sum of squared observations by the number of observations for each sire. 

The expression $y^TZ$ is a vector with sums of observation for each sire. 


```{r}
crossprod(mat_z_p02)
```

Hence the complete expression is

```{r}
ytz %*%solve(crossprod(mat_z_p02)) %*% t(ytz)
```

```{r}
ytz
```

```{r}
ytz %*% crossprod(mat_z_p02)
```


The same result should be obtained by

```{r}
ytz^2 / diag(crossprod(mat_z_p02))
```

```{r}
sum(ytz^2 / diag(crossprod(mat_z_p02)))
```


```{r}
ytz
(ytz / diag(crossprod(mat_z_p02)))
```

```{r}
diag(crossprod(mat_z_p02))
```

