---
title: "Solution To Exercise Converter"
author: "Peter von Rohr"
date: "2018-11-07"
output: html_notebook
---

## Disclaimer
The experiments described in this document are implemented in package `teachtools` in the function `convert_sol_to_ex()`.


## Backgroud
This notebook contains a few experiments with automated conversion of a solution document into an exercise document. The approach used here builds on the assumption that all solution documents are Rmarkdown (.Rmd) documents and have the same structure. This assumption can be satisfied using unique templates for the solution documents. 


## Document Structure
The following block shows the assumed structure of a solution document

```
---
<YAML-Header>
---

## Problem 1: <Title of Problem 1>

### Solution

## Problem 2: <Title of Problem 2>

### Solution

...
```


## Identification of Blocks
From this structure, we can see that Problem sections always start with `## Problem` and solution sections start with `### Solution`. Hence when reading a solutions document into a character vector, we can search for these starting tags of the different sections. This is shown in the following example.

```{r}
### # try to get path to solution document
s_proj_path <- rprojroot::find_rstudio_root_file()
### # read a solutions document
s_sol_doc <- "sol/w08/lbg_fs_2018_w08_sol07.Rmd"
s_sol_path <- file.path(s_proj_path, s_sol_doc)
### # check that s_sol_doc exists
stopifnot(file.exists(s_sol_path))
con_sol_doc <- file(s_sol_path)
vec_sol <- readLines(con = con_sol_doc)
### # check its header
head(vec_sol)
close(con_sol_doc)
```

Now that we have read the solutions document, we do the actual search

```{r}
### # define patterns
s_prob_pat <- '## Problem'
s_sol_pat <- '### Solution'
(vec_prob_start <- grep(s_prob_pat, vec_sol, fixed = TRUE))
(vec_sol_start <- grep(s_sol_pat, vec_sol, fixed = TRUE))
### # consistency, the number of blocks must be equal
stopifnot(length(vec_prob_start) == length(vec_sol_start))
```

The above output shows us that the first problem section starts at line `r vec_prob_start[1]` and it ends at line `r vec_sol_start[1]-1`. The first solution section starts at line `r vec_sol_start[1]` and it ends at line `r vec_prob_start[2]-1`. 


## Extracting Exercise Blocks
The exercise blocks that we want to extract are between line 1 and `r vec_sol_start[1]-1]` and `r vec_prob_start[2]` and `r vec_sol_start[2]-1`. We are putting these sections together using a simple `c()` over all problems. Initially, we modify the title of the solution document to fit the title of an exercise.

```{r}
### # replace Solution
s_sol_title <- 'Solution'
s_ex_title <- 'Exercise'
vec_ex <- gsub(s_sol_title, s_ex_title, vec_sol[1:(vec_prob_start[1]-1)], fixed = TRUE)
for (lidx in seq_along(vec_prob_start)){
  cat("[INFO -- ] Line index: ", lidx, "\n")
  ### # assign or concatenate
  if (is.null(vec_ex)){
    vec_ex <- vec_sol[vec_prob_start[lidx]:(vec_sol_start[lidx]-1)]
  } else {
    vec_ex <- c(vec_ex, vec_sol[vec_prob_start[lidx]:(vec_sol_start[lidx]-1)])
  }
}
head(vec_ex)
```

Once we have done the extraction, we can write the exercise to a file. 

```{r}
### # generate directory and file name for ex
s_ex_doc <- gsub('sol', 'ex', s_sol_doc, fixed = TRUE)
s_ex_path <- file.path(s_proj_path, s_ex_doc)
s_ex_dir <- dirname(s_ex_path)
### # check whether s_ex_dir exists, create if not
if (!dir.exists(s_ex_dir)) {
  cat("[INFO -- ] Create directory: ", s_ex_dir, "\n")
  dir.create(s_ex_dir)
}
### # write vec_ex to s_ex_path
cat("[INFO -- ] Writing vec_ex to: ", s_ex_path, "\n")
cat(paste(vec_ex, collapse  = "\n"), "\n", file = s_ex_path)
```



