---
title: "Generate Data for Problem 2 in Exercise 11"
author: "Peter von Rohr"
date: "`r Sys.Date()` (`r Sys.info()[['user']]`)"
output: html_notebook
---


## Disclaimer
This notebook shows how data for problem 2 in exercise 11 are generated.


## The Model
The animal model is used to generate the data. 

$$y = X\beta + Za + e$$


```{r datasetup, echo=FALSE, results='hide'}
set.seed(4321)
### # fixed effects
n_nr_obs <- 5
n_nr_fixed_effects <- 1
n_mu <- 13.41
mat_x <- matrix(1, nrow = n_nr_obs, ncol = n_nr_fixed_effects)

### # variance components
n_sigmaa2 <- 8
n_sigmae2 <- 24

### # pedigree
tbl_pedi_ex11prob02 <- tibble::data_frame(Animal = (1:n_nr_obs),
                    Sire = c(NA,NA,NA,1,4),
                    Dam = c(NA,NA,NA,2,3))

### # assuming all animals have an observation, we get the matrix z
### # as identity matrix
mat_z <- diag(nrow(tbl_pedi_ex11prob02))
```

The following function was used in `20181111_generate_data_ex8p2.Rmd`. 

```{r loopfunctionbv}
### # function to get breeding values
#' @title Generate vector of breeding values 
#'
#' @description
#' The breeding values are generated using a loop
#' over scalar values
generate_bv <- function(ptbl_ped, pn_sigmaa2){
  ### # get matrix D based on pedigree
  ped <- pedigreemm::pedigree(sire = ptbl_ped$Sire, dam = ptbl_ped$Dam, label = as.character(ptbl_ped$Animal))
  diag_mat_d <- as.matrix(pedigreemm::Dmat(ped = ped))
  
  ### # number of animals in pedigree
  n_nr_ani_ped <- nrow(ptbl_ped)
  ### # intialize a result vector
  vec_bv_result <- rep(0, n_nr_ani_ped)
  ### # loop over animals in pedigree and generate the breeding values
  for (idx in seq_along(ptbl_ped$Animal)){
    ### # get parent breeding values
    sire_bv <- 0
    if (!is.na(ptbl_ped$Sire[idx]))
      sire_bv <- vec_bv_result[ptbl_ped$Sire[idx]]
    dam_bv <- 0
    if (!is.na(ptbl_ped$Dam[idx]))
      dam_bv <- vec_bv_result[ptbl_ped$Dam[idx]]
    ### # sample mendelian sampling effect
    n_mend_sampl <- rnorm(1, mean = 0, sd = sqrt(diag_mat_d[idx,1] * pn_sigmaa2))
    vec_bv_result[idx] <- 0.5 * sire_bv + 0.5 * dam_bv + n_mend_sampl
  }
  ### # return result vector
  return(vec_bv_result)
}

```

We want to vectorize the above function. The core process is the generation of the vector $a$ of breeding values using the $P$ matrix in the simple decomposition

$$a = P \cdot a + m$$

The matrix $P$ can be computed by solving both decompositions for $m$ and getting $P$ as

$$a = L \cdot m$$

From this we get, 

$$P = I - L^{-1}$$

The matrix $L$ is computed based on the cholesky decomposition of $A$ given by

$$A = R \cdot R^T$$

Setting $R = L \cdot S$ with $D = S \cdot S$ and $A = L \cdot D \cdot L^T$. The following steps can be used

* From the pedigree, we get $A$
* $R$ is obained by `chol(A)`
* $D$ is obtained by `pedigreemm::Dmat(ped)`
* $S$ is obtained by `sqrt(D)`
* $L$ is computed as $L = R \cdot S^{-1}$
* $P$ is obtained as $P = I - L^{-1}$


```{r vecfunctionbv}
### # function to get breeding values
#' @title Generate vector of breeding values 
#'
#' @description
#' Breeding values are generated using a vectorized 
#' version of the generation of breeding values
generate_vec_bv <- function(ptbl_ped, pn_sigmaa2){
  ### # get pedigree
  ped <- pedigreemm::pedigree(sire = ptbl_ped$Sire, dam = ptbl_ped$Dam, label = as.character(ptbl_ped$Animal))
  ### # number of animal
  n_nr_ani <- nrow(ptbl_ped)
  ### # get matrix D
  diag_mat_d <- diag(pedigreemm::Dmat(ped = ped), nrow = n_nr_ani, ncol = n_nr_ani)
  ### # get matrix A based on pedigree
  mat_a <- as.matrix(pedigreemm::getA(ped = ped))
  ### # cholesky of A
  mat_r <- t(chol(mat_a))
  ### # sqrt(D) to mat_s
  mat_s <- sqrt(diag_mat_d)
  ### # matrix L
  mat_l <- mat_r %*% solve(mat_s)
  ### # finally get matrix P
  mat_p <- diag(n_nr_ani) - solve(mat_l)
  ### # sample the vector of mendelian sampling
  vec_a <- rnorm(n_nr_ani, mean = 0, sd = sqrt(diag(diag_mat_d) * pn_sigmaa2))
  ### # adding pedigree Information
  vec_a_result <- mat_p %*% vec_a + vec_a
  ### # return result
  return(vec_a_result)
}
```

Computing the breeding values with the old function leads to

```{r}
# generate_bv(ptbl_ped = tbl_pedi_ex11prob02, pn_sigmaa2 = n_sigmaa2)
# -1.2070522 -0.6324697  2.0296985  0.7631304  1.1396999
```

Use the new function

```{r}
vec_a <- generate_vec_bv(ptbl_ped = tbl_pedi_ex11prob02, pn_sigmaa2 = n_sigmaa2)
```


## Fixed Effects
The fixed effects are reduced to a general mean $\mu$. 


## Observations
Observations are generated based on the given model as

```{r}
vec_y <- crossprod(t(mat_x), n_mu) + crossprod(t(mat_z), vec_a) + rnorm(n_nr_obs, mean = 0, sd = sqrt(n_sigmae2))
```

Add observations to pedigree and write data to a data file

```{r, message=FALSE, warning=FALSE}
library(dplyr)
tbl_data_ex11p02 <- tbl_pedi_ex11prob02 %>%
  mutate(Observation = round(vec_y, digits = 2))
readr::write_csv(tbl_data_ex11p02, path = "data_ex11_p02.csv")
```

## Sire Relationship Matrix

In Mrode 2005, the inverse of a sire relationship matrix is given as

```{r}
Asireinv = matrix(c(1.3333, 0, -.6667,
                    0, 1, 0,
                    -.6667, 0, 1.333), nrow = 3, byrow = TRUE)
Asireinv
```

The relationship matrix is

```{r}
solve(Asireinv)
```

