---
title: Livestock Breeding and Genomics - Exercise 12
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

```{r TaskCounts, echo=FALSE}
cnt <- rmddochelper::R6ClassCount$new()
cnt$set_prefix(ps_prefix = "## Problem")
```

```{r, echo=FALSE, results='asis'}
cat(cnt$out(ps_suffix = "Multivariate BLUP Animal Model"), "\n")
```

```{r sol12p01data, echo=FALSE, results='hide'}
n_nr_trait <- 2
n_nr_founder <- 3
n_nr_animal <- 8
n_nr_observation <- n_nr_animal - n_nr_founder
tbl_data_sol12p01 <- tibble::data_frame(Animal = c((n_nr_founder+1):n_nr_animal),
                                        Sex = c("Male", "Female","Female","Male","Male"),
                                        Sire = c(1,3,1,4,3),
                                        Dam = c(NA,2,2,5,6),
                                        WWG = c(4.5,2.9,3.9,3.5,5.0),
                                        PWG = c(6.8,5.0,6.8,6.0,7.5))
```

The table below contains data for pre-weaning gain (WWG) and post-weaning gain (PWG) for `r n_nr_observation` beef calves.

```{r sol12p01table, echo=FALSE, results='asis'}
knitr::kable(tbl_data_sol12p01,
             booktabs = TRUE,
             longtable = TRUE)
```

The genetic variance-covariance matrix $G_0$ between the traits is 

```{r}
mat_g0 <- matrix(data = c(20,18,18,40), nrow = n_nr_trait, byrow = TRUE)
cat("$$\n")
cat("G_0 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_g0, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The residual variance-covariance matrix $R_0$ between the traits is

```{r}
mat_r0 <- matrix(data = c(40,11,11,30), nrow = n_nr_trait, byrow = TRUE)
cat("$$\n")
cat("R_0 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_r0, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


### Your Task
Set up the mixed model equations for a multivariate BLUP analysis and compute the estimates for the fixed effects and the predictions for the breeding values.



\vspace{3ex}

```{r, echo=FALSE, results='asis'}
cat(cnt$out(ps_suffix = "Variance Components Estimation"), "\n")
```
The simplest forM of variance components estimation is based on the residuals of a fitted linear model and is shown in the summary results of the R-function `lm()`. Let us assume that we are given the dataset in the table shown below to which we fit a simple sire model. 


```{r datasol12p02}
tbl_data_sol12p02 <- tibble::data_frame( Animal = c(4, 5, 6, 7, 8),
                                         Sire   = c(2, 1, 1, 2, 1),
                                         WWG    = c(4.5,2.9,3.9,3.5,5.0) )

knitr::kable(tbl_data_sol12p02,
             booktabs  = TRUE,
             longtable = TRUE,
             caption   = "Example Dataset for Variance Components Estimation Based on Residuals Using a Sire Model")

```

The sire model is simplified to have a common mean $\mu$.  For a moment we are setting the sire effects to be fixed effects. This leads to the following model with $var(e) = I * \sigma_e^2$

$$y = X\mu + Z_ss + e$$

Using the above shown dataset we can use the R-function `lm()` to fit this simple linear model. Because, we want to have the sires as fixed effects, we have to convert them into factors before calling `lm()`.

```{r, echo=TRUE, results='markup'}
tbl_data_sol12p02$Sire <- as.factor(tbl_data_sol12p02$Sire)
lm_data_sol12p02 <- lm( WWG ~ 1 + Sire, data = tbl_data_sol12p02 )
summary(lm_data_sol12p02)
```

```{r, echo=FALSE, results='hide'}
res_std_sol12p02 <- sigma(lm_data_sol12p02)
```

From the output of `summary()` we are given the residual standard error to be `r round(res_std_sol12p02, digits = 4)`. This residual standard error is an estimate of $\sigma_e$. The question is where does it come from. The least-squares procedure does not yield this estimate for $\sigma_e$. The answer is that this estimate comes from the residuals $r$ of the model. For our model the vector $r$ of residuals is defined as

$$r = y - X\widehat{\mu} - Z_s\widehat{s}$$

where $\widehat{\mu}$ and $\widehat{s}$ can be taken from the ouput of the `summary()` function. They correspond to

```{r, echo=FALSE, results='asis'}
vec_coeff <- coefficients(lm_data_sol12p02)
n_mu_hat <- vec_coeff[[1]]
vec_sire_hat <- c(0, vec_coeff[[2]])

cat("$$\\widehat{\\mu} = ", n_mu_hat, "$$\n")
cat("$$\n")
cat("\\widehat{s} = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_sire_hat, pnDigits = 4), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The estimate $\widehat{\sigma_e^2}$ for $\sigma_e^2$ is obtained by

$$\widehat{\sigma_e^2} = \frac{1}{n-p} \sum_{i=1}^n r_i^2$$

where $n$ is the total number of observations and $p$ is the number of parameters that are estimated by `lm` which is `r length(vec_coeff)` for our sire model. The term $n-p$ is also called __degrees of freedom__ (df). What is given as residual standard error by the output of `summary()` is the square root of $\widehat{\sigma_e^2}$.

### Your Task
Verify for the above given dataset and the proposed sire model the residual standard error given by `summary()` by using the computation based on the residuals shown above.

