---
title: "R Notebook"
output: html_notebook
---


## Number of Sires in a Pedigree

```{r}
tbl_ped_exmp3 <- tibble::data_frame(Animal = c(4, 5, 6, 7),
                                    Sire   = c(1, 1, 4, 6),
                                    Dam    = c(2, 3, 2, 5))
(n_nr_sire <- nlevels(as.factor(tbl_ped_exmp3$Sire)))
```

## Find unique sire ids
```{r}
vec_sire_idx <- unique(tbl_ped_exmp3$Sire)
paste0("s_", vec_sire_idx, collapse = "")
sapply(vec_sire_idx, function(x) paste0("s_{", x, "}"), USE.NAMES = FALSE)
```

```{r}
n_nr_obs <- 4
### # intercept and two levels of a fixed effect
vec_beta <- c(10.34, 5.2, -1.92)
mat_x <- matrix(c(1, 1, 0,
                  1, 0, 1,
                  1, 0, 1,
                  1, 0, 1), nrow = n_nr_obs, byrow = TRUE)
mat_x
```

```{r}
mat_x[,-1]
```


### Progeny Performance

```{r BvPredData, echo=FALSE, results='hide'}
set.seed(1521)
n_nr_obs <- 4
### # intercept and two levels of a fixed effect
vec_beta <- c(10.34, 5.2, -1.92)
mat_x <- matrix(c(1, 1, 0,
                  1, 0, 1,
                  1, 0, 1,
                  1, 0, 1), nrow = n_nr_obs, byrow = TRUE)
### # find number of herds, knowing that first column is intercept
n_nr_herd <- ncol(mat_x)-1
### # variance components
n_sigmae2 <- 60
n_sigmaa2 <- 20
n_sigmas2 <- n_sigmaa2 / 4
n_h2 <- n_sigmaa2 / (n_sigmaa2 + n_sigmae2)

### # pedigree
tbl_ped_exmp3 <- tibble::data_frame(Animal = c(4, 5, 6, 7),
                                    Sire   = c(1, 1, 3, 6),
                                    Dam    = c(2, 4, 2, 5))
### # number of animals, number of founders and number of sirees
n_nr_ani_exmp3pedext <- max(tbl_ped_exmp3$Animal)
n_nr_founder_exmp3pedext <- min(tbl_ped_exmp3$Animal) - 1
n_nr_sire_exmp3pedext <- nlevels(as.factor(tbl_ped_exmp3$Sire))
### # extend pedigree with founders
tbl_ped_exmp3_ext <- tibble::data_frame(Animal = c(1:n_nr_ani_exmp3pedext),
                                        Sire   = c(rep(NA, n_nr_founder_exmp3pedext), tbl_ped_exmp3$Sire),
                                        Dam    = c(rep(NA, n_nr_founder_exmp3pedext), tbl_ped_exmp3$Dam))

### # breeding values
### # function to get breeding values
#' @title Generate vector of breeding values 
#'
#' @description
#' Breeding values are generated using a vectorized 
#' version of the generation of breeding values
generate_vec_bv <- function(ptbl_ped, pn_sigmaa2){
  ### # get pedigree
  ped <- pedigreemm::pedigree(sire = ptbl_ped$Sire, dam = ptbl_ped$Dam, label = as.character(ptbl_ped$Animal))
  ### # number of animal
  n_nr_ani <- nrow(ptbl_ped)
  ### # get matrix D
  diag_mat_d <- diag(pedigreemm::Dmat(ped = ped), nrow = n_nr_ani, ncol = n_nr_ani)
  ### # get matrix A based on pedigree
  mat_a <- as.matrix(pedigreemm::getA(ped = ped))
  ### # cholesky of A
  mat_r <- t(chol(mat_a))
  ### # sqrt(D) to mat_s
  mat_s <- sqrt(diag_mat_d)
  ### # matrix L
  mat_l <- mat_r %*% solve(mat_s)
  ### # finally get matrix P
  mat_p <- diag(n_nr_ani) - solve(mat_l)
  ### # sample the vector of mendelian sampling
  vec_a <- rnorm(n_nr_ani, mean = 0, sd = sqrt(diag(diag_mat_d) * pn_sigmaa2))
  ### # adding pedigree Information
  vec_a_result <- mat_p %*% vec_a + vec_a
  ### # return result
  return(vec_a_result)
}
### # vector of breeding values
vec_bv <- generate_vec_bv(ptbl_ped = tbl_ped_exmp3_ext, pn_sigmaa2 = n_sigmaa2)
### # design matrix Z
mat_z <- cbind(matrix(0, nrow = n_nr_obs, ncol = n_nr_founder_exmp3pedext), diag(1, nrow = n_nr_obs, ncol = n_nr_obs))

### # generate observations
vec_y <- crossprod(t(mat_x), vec_beta) + crossprod(t(mat_z), vec_bv) + rnorm(n_nr_obs, mean = 0, sd = sqrt(n_sigmae2))

### # population mean
n_mu <- mean(vec_y)

library(dplyr)
mat_herd <- crossprod(t(mat_x), c(0,1,2))
tbl_dat_exmp3 <- tbl_ped_exmp3 %>% mutate(Herd = mat_herd[,1], Observation = round(vec_y, digits = 2))
knitr::kable(tbl_dat_exmp3, booktabs = TRUE, longtable = TRUE)

```

Number of progeny per sire

```{r}
(tbl_dat_exmp3_nrprog <- tbl_dat_exmp3 %>%
  group_by(Sire) %>%
  summarise(`Number of Progeny` = n()))
```


Compute mean observations for offspring per parent

```{r}
(tbl_dat_exmp3_meanobs <- tbl_dat_exmp3 %>%
  group_by(Sire) %>%
  summarise(`Mean Observation` = mean(Observation)))
```


Join the two tibbles

```{r}
 (tbl_dat_exmp3_meanobs <- tbl_dat_exmp3_meanobs %>%
  inner_join(tbl_dat_exmp3_nrprog))
```

Get breeding values

```{r}

```


```{r}
 (tbl_dat_exmp3 <- tbl_dat_exmp3 %>%
  inner_join(tbl_dat_exmp3_nrprog))
```



## Problem 5: Pedigree

The original version of the pedigree table

```{r, echo=FALSE, results='asis'}
tbl_ped_exmp5_reord <- tibble::data_frame(Animal = c("SIRA", "Flurina", "RENARD", "FREUDI"),
                                    Sire   = c("TELL", "HARRY", "LARS", "RENARD"),
                                    Dam    = c("SILVA", "Freia", "SIRA", "Flurina"))

knitr::kable(tbl_ped_exmp5_reord, booktabs = TRUE, longtable = TRUE)
```

Recoding the pedigree to numers with the following mapping

```{r, echo=FALSE, results='asis'}
tbl_ped_exmp5_map <- tibble::data_frame(Name = c("TELL","SILVA","HARRY","Freia","LARS","SIRA","Flurina", "RENARD", "FREUDI"),
                                        Code = c(1:9))
knitr::kable(tbl_ped_exmp5_map, booktabs = TRUE, longtable = TRUE)
```

Applying the above shown mapping

```{r, echo=FALSE, results='asis'}
tbl_ped_exmp5_recode <- tibble::data_frame(Animal = c(6, 7, 8, 9),
                                    Sire   = c(1, 3, 5, 8),
                                    Dam    = c(2, 4, 6, 7))
knitr::kable(tbl_ped_exmp5_recode, booktabs = TRUE, longtable = TRUE)
```

Extending the pedigree with founders

```{r, echo=FALSE, results='asis'}
n_nr_founder_emp5 <- min(tbl_ped_exmp5_recode$Animal)-1
n_nr_ani_emp5 <- max(tbl_ped_exmp5_recode$Animal)
tbl_ped_exmp5_ext <- tibble::data_frame(Animal = c(1:n_nr_ani_emp5),
                                        Sire   = c(rep(NA, n_nr_founder_emp5), tbl_ped_exmp5_recode$Sire),
                                        Dam    = c(rep(NA, n_nr_founder_emp5), tbl_ped_exmp5_recode$Dam))
knitr::kable(tbl_ped_exmp5_ext, booktabs = TRUE, longtable = TRUE)
```

This pedigree can be used to compute the numerator relationship matrix

```{r, echo=FALSE, results='asis'}
ped_exmp5 <- pedigreemm::pedigree(sire = tbl_ped_exmp5_ext$Sire,
                                  dam  = tbl_ped_exmp5_ext$Dam,
                                  label = as.character(tbl_ped_exmp5_ext$Animal))
mat_a_exmp5 <- as.matrix(pedigreemm::getA(ped = ped_exmp5))
mat_ainv_exmp5 <- as.matrix(pedigreemm::getAInv(ped = ped_exmp5))


# cat("$$\n")
# cat("A = \\left[ \n")
# cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_a_exmp5, pnDigits = 3), collapse = "\n"), "\n")
# cat("\\right] \n")
# cat("$$\n")
```


Use row and column numbers to be able to specify an order

```{r}
colnames(mat_a_exmp5) <- tolower(tbl_ped_exmp5_map$Name)
rownames(mat_a_exmp5) <- tolower(tbl_ped_exmp5_map$Name)
colnames(mat_ainv_exmp5) <- tolower(tbl_ped_exmp5_map$Name)
rownames(mat_ainv_exmp5) <- tolower(tbl_ped_exmp5_map$Name)
 mat_a_exmp5
```

Define a different order

```{r}
# vec_name <- c("lars", "harry", "freia", "tell", "silva", "sira", "renard", "flurina", "freudi")
# vec_name <- c("freia","silva","harry","tell","lars","flurina","sira","renard","freudi")
# vec_name <- c("tell", "silva", "sira", "harry", "freia","flurina","lars","renard","freudi")
vec_name <- c("tell", "silva", "harry", "freia", "lars", "sira", "flurina", "renard", "freudi")
mat_a_exmp5[vec_name, vec_name]
```


```{r}
mat_ainv_exmp5[vec_name, vec_name]
```

