---
title: "Inverse Numerator Relationship Matrix"
author: "Peter von Rohr"
date: "16 November 2018"
output:
  beamer_presentation:
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "asis")
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Structure of $A^{-1}$

* Look at a simple example of $A$ and $A^{-1}$

```{r simplepedexamplesetup, echo=FALSE, results='hide'}
### # first six animals from Goetz p. 83
suppressPackageStartupMessages( library(pedigreemm) )
n_nr_ani_ped <- 5
n_nr_parent <- 3
tbl_ped_simple <- dplyr::data_frame(Calf = c(1:n_nr_ani_ped),
                             Sire = c(NA, NA, NA, 1, 3),
                             Dam  = c(NA, NA, NA, 2, 2))
### # pedigreemm
ped_simple <- pedigree(sire = tbl_ped_simple$Sire, dam = tbl_ped_simple$Dam, label = as.character(1:n_nr_ani_ped))
matA_simple <- as.matrix(getA(ped = ped_simple))
matAinv_simple <- as.matrix(getAInv(ped = ped_simple))
### # LDL decomposition based on cholesky
matR <- t(chol(matA_simple))
### # matS = sqrt(matD)
matD <- diag(Dmat(ped = ped_simple), n_nr_ani_ped)
matS <- sqrt(matD)
matL <- matR %*% solve(matS)
```

```{r tabpedsimpleexample, echo=FALSE}
knitr::kable(tbl_ped_simple,
             booktabs = TRUE,
             longtable = TRUE,
             caption = "Pedigree Used To Compute Inverse Numerator Relationship Matrix")
```


## Numerator Relationship Matrix $A$

```{r displaymatrixasimple, echo=FALSE, results='asis'}
cat("\\begin{equation}\n")
cat("A = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matA_simple, pnDigits = 4), sep = "\n"), "\n")
cat("\\right]\n")
cat("\\label{eq:numeratorrelationshipmatrix}\n")
cat("\\end{equation}\n")
```


## Inverse Numerator Relationship Matrix $A^{-1}$

```{r displaymatrixainv, echo=FALSE, results='asis'}
cat("$$\n")
cat("A^{-1} = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matAinv_simple, pnDigits = 4), sep = "\n"), "\n")
cat("\\right]")
cat("$$")
```


## Conclusions

* $A^{-1}$ has simpler structure than $A$ itself
* Non-zero elements only at positions of parent-progeny and parent-mate positions
* Parent-mate positions are positive, parent-progeny are negative


## Henderson's Rules

* Based on `LDL`-decomposition of $A$

$$A= L * D * L^T$$

\begin{tabular}{lll}
where  &  $L$  & Lower triangular matrix \\
       &  $D$  & Diagonal matrix
\end{tabular}

* Why?
    + matrices $L$ and $D$ can be inverted directly, we 'll see how ...
    + construct $A^{-1} = (L^T)^{-1} * D^{-1} * L^{-1}$
    

## Example

```{r examplematlmatd, echo=FALSE, results='asis'}
cat("$$\n")
cat("L = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matL, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
cat("$$\n")
cat("D = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matD, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

$\rightarrow$ Verify that $A = L * D * L^T$


## Decomposition of True Breeding Value

* True breeding value ($a_i$) of animal $i$

$$
a_i = {1\over 2} a_s + {1\over 2} a_d + m_i
$$

* Do that for all animals in pedigree


## Decomposition for Example

\begin{align}
a_1  &=  m_1 \notag \\
a_2  &=  m_2 \notag \\
a_3  &=  m_3 \notag \\
a_4  &=  {1\over 2} a_1 + {1\over 2} a_2 + m_4 \notag \\
a_5  &=  {1\over 2} a_3 + {1\over 2} a_2 + m_5 \notag
\end{align}


## Matrix Vector Notation

* Define vectors $a$ and $m$ as
* Coefficients of $a_s$ and $a_d$ into matrix $P$

```{r, echo=FALSE, results='asis'}
### # definition of vectors a and m
vec_a <- rmddochelper::vecGetVecElem(psBaseElement = "a", pnVecLen = n_nr_ani_ped, psResult = "latex")
vec_m <- rmddochelper::vecGetVecElem(psBaseElement = "m", pnVecLen = n_nr_ani_ped, psResult = "latex")
### # define matrix P
matP <- matrix(0, nrow = n_nr_ani_ped, ncol = n_nr_ani_ped)
matP[4,1] <- matP[4,2] <- matP[5,3] <- matP[5,2] <- 0.5
### # show both vectors
cat("$$\n")
cat("a = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a), collapse = "\n"), "\n")
cat("\\right] \\text{, }\n")
cat("m = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_m), collapse = "\n"), "\n")
cat("\\right] \\text{, }\n")
cat("P = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matP, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

* Result: Decomposition of true breeding values

$$a = P \cdot a + m$$


## Decomposition of Variance

* Analogous decomposition of $var(a_i)$

\begin{align}
var(a_i)  &=  var(1/2a_s + 1/2a_d + m_i) \notag \\
          &=  var(1/2a_s)  + var(1/2a_d) + {1\over 2} * cov(a_s, a_d) + var(m_i) \notag \\
          &=  1/4 var(a_s) + 1/4 var(a_d) + {1\over 2} * cov(a_s, a_d) + var(m_i) \notag 
\end{align}

* From the definition of $A$

\begin{align}
var(a_i)  &=  (1 + F_i) \sigma_a^2 \notag \\
var(a_s)  &=  (1 + F_s) \sigma_a^2 \notag \\
var(a_d)  &=  (1 + F_d) \sigma_a^2 \notag \\
cov(a_s, a_d)  &=  (A)_{sd} \sigma_a^2 = 2F_i \sigma_a^2 \notag
\end{align}


## Variance of Mendelian Sampling Terms

* What is $var(m_i)$?

* Solve equation for $var(a_i)$ for $var(m_i)$

\begin{align}
var(m_i)  &=  var(a_i) - 1/4 var(a_s) - 1/4 var(a_d) - 2 * cov(a_s, a_d) \notag
\end{align}

* Insert definitions from $A$

\begin{align}
var(m_i)  &=  (1 + F_i) \sigma_a^2 - 1/4 (1 + F_s) \sigma_a^2 - 1/4 (1 + F_d) \sigma_a^2 - {1\over 2} * 2 * F_i \sigma_a^2 \notag \\
          &=  \left({1\over 2} - {1\over 4}(F_s + F_d)\right) \sigma_a^2 \notag
\end{align}

* True, for both parents $s$ and $d$ of animal $i$ are known


## Unknown Parents

* Only parent $s$ of animal $i$ is known

\begin{align}
a_i       &=  {1\over 2} a_s + m_i \notag \\
var(m_i)  &=  \left(1 - {1\over 4}(1+ F_s)\right) \sigma_a^2 \notag \\
          &=  \left({3\over 4} - {1\over 4}F_s\right) \sigma_a^2 \notag
\end{align}

* Both parents are unknown

\begin{align}
a_i       &=  m_i \notag \\
var(m_i)  &=  \sigma_a^2 \notag
\end{align}


## Recursive Decomposition

* True breeding values of $s$ and $d$ can be decomposed into

\begin{align}
a_s  &=  {1\over 2} a_{ss} + {1\over 2} a_{ds} + m_s \notag \\
a_d  &=  {1\over 2} a_{sd} + {1\over 2} a_{dd} + m_d \notag
\end{align}

\begin{tabular}{lll}
where  &  $ss$  &  sire of $s$  \\
       &  $ds$  &  dam of $s$   \\
       &  $sd$  &  sire of $d$  \\
       &  $dd$  &  dam of $d$  
\end{tabular}


## Example

* Add animal 6 with parents 4 and 5 to our example pedigree

```{r showextpedigree, echo=FALSE, results='asis'}
n_nr_ani_ped <- 6
n_nr_parent <- 3
tbl_ped_ext <- dplyr::data_frame(Calf = c(1:n_nr_ani_ped),
                             Sire = c(NA, NA, NA, 1, 3, 4),
                             Dam  = c(NA, NA, NA, 2, 2, 5))
### # pedigreemm
ped_ext <- pedigree(sire = tbl_ped_ext$Sire, dam = tbl_ped_ext$Dam, label = as.character(1:n_nr_ani_ped))
matA_ext <- as.matrix(getA(ped = ped_ext))
matAinv_ext <- as.matrix(getAInv(ped = ped_ext))
### # LDL decomposition based on cholesky
matR_ext <- t(chol(matA_ext))
### # matS = sqrt(matD)
matD_ext <- diag(Dmat(ped = ped_ext), n_nr_ani_ped)
matS_ext <- sqrt(matD_ext)
matL_ext <- matR_ext %*% solve(matS_ext)
### # show table
knitr::kable(tbl_ped_ext, 
             booktabs  = TRUE,
             longtab   = TRUE)
```


## First Step Of Decomposition

\begin{align}
a_1  &=  m_1 \notag \\
a_2  &=  m_2 \notag \\
a_3  &=  m_3 \notag \\
a_4  &=  {1\over 2} a_1 + {1\over 2} a_2 + m_4 \notag \\
a_5  &=  {1\over 2} a_3 + {1\over 2} a_2 + m_5 \notag \\
a_6  &=  {1\over 2} a_4 + {1\over 2} a_5 + m_6 \notag
\end{align}


## Decompose Parents

\begin{align}
a_1  &=  m_1 \notag \\
a_2  &=  m_2 \notag \\
a_3  &=  m_3 \notag \\
a_4  &=  {1\over 2} m_1 + {1\over 2} m_2 + m_4 \notag \\
a_5  &=  {1\over 2} m_3 + {1\over 2} m_2 + m_5 \notag \\
a_6  &=  {1\over 2} \left({1\over 2}(a_1 + a_2) + m_4\right) + {1\over 2} \left({1\over 2} (a_3 + a_2) + m_5 \right) + m_6 \notag \\
     &=  {1\over 4} (a_1 + a_2) + {1\over 2} m_4 + {1\over 4} (a_3 + a_2) + {1\over 2} m_5 + m_6 \notag
\end{align}


## Decompose Grand Parents

* Only animal 6 has true breeding values for grand parents

\begin{align}
a_6  &=   {1\over 4} (a_1 + a_2) + {1\over 2} m_4 + {1\over 4} (a_3 + a_2) + {1\over 2} m_5 + m_6 \notag \\
     &=   {1\over 4} m_1 + {1\over 4} m_2 + {1\over 4} m_3 + {1\over 4} m_2 + {1\over 2} m_4 + {1\over 2} m_5 + m_6 \notag \\
     &=   {1\over 4} m_1 + {1\over 2} m_2 + {1\over 4} m_3 + {1\over 2} m_4 + {1\over 2} m_5 + m_6 \notag
\end{align}


## Summary

\begin{align}
a_1  &=  m_1 \notag \\
a_2  &=  m_2 \notag \\
a_3  &=  m_3 \notag \\
a_4  &=  {1\over 2} m_1 + {1\over 2} m_2 + m_4 \notag \\
a_5  &=  {1\over 2} m_3 + {1\over 2} m_2 + m_5 \notag \\
a_6  &=   {1\over 4} m_1 + {1\over 2} m_2 + {1\over 4} m_3 + {1\over 2} m_4 + {1\over 2} m_5 + m_6 \notag
\end{align}


## Matrix-Vector Notation

* Use vectors $a$ and $m$ again

```{r, echo=FALSE, results='asis'}
### # definition of vectors a and m
vec_a_ext <- rmddochelper::vecGetVecElem(psBaseElement = "a", pnVecLen = n_nr_ani_ped, psResult = "latex")
vec_m_ext <- rmddochelper::vecGetVecElem(psBaseElement = "m", pnVecLen = n_nr_ani_ped, psResult = "latex")
### # show both vectors
cat("\\small \n $$\n")
cat("a = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a_ext), collapse = "\n"), "\n")
cat("\\right] \\text{, }\n")
cat("m = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_m_ext), collapse = "\n"), "\n")
cat("\\right] \\text{, }\n")
cat("L = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matL_ext, pnDigits = 2), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n \\normalsize \n")
```

* Result of recursive decomposition of $a_i$

$$a = L \cdot m$$

## Variance From Recursive Decomposition

\begin{align}
var(a)  &=  var(L \cdot m) \notag \\
        &=  L \cdot var(m) \cdot L^T \notag
\end{align}

where $var(m)$ is the variance-covariance matrix of all components in vector $m$. 

* covariances of components $m_i$, $cov(m_i, m_j) = 0$ for $i \ne j$
* $var(m_i)$ computed as shown before


## Result

* variance-covariance matrix $var(m)$ can be written as $D * \sigma_a^2$ where $D$ is diagnoal

\begin{align}
\rightarrow var(a)  &= L \cdot var(m) \cdot L^T \notag \\
                    &= L \cdot D * \sigma_a^2 \cdot L^T \notag \\
                    &= L \cdot D\cdot L^T  * \sigma_a^2  \notag \\
                    &= A \sigma_a^2 \notag
\end{align}

\begin{align}
\rightarrow A  &= L \cdot D\cdot L^T \notag
\end{align}
