---
title: Livestock Breeding and Genomics - Solution 12
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', fig.pos = 'H')
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

```{r TaskCounts, echo=FALSE}
cnt <- rmddochelper::R6ClassCount$new()
cnt$set_prefix(ps_prefix = "## Problem")
```

```{r, echo=FALSE, results='asis'}
cat(cnt$out(ps_suffix = "Multivariate BLUP Animal Model"), "\n")
```

```{r sol12p01data, echo=FALSE, results='hide'}
n_nr_trait <- 2
n_nr_founder <- 3
n_nr_animal <- 8
n_nr_observation <- n_nr_animal - n_nr_founder
tbl_data_sol12p01 <- tibble::data_frame(Animal = c((n_nr_founder+1):n_nr_animal),
                                        Sex = c("Male", "Female","Female","Male","Male"),
                                        Sire = c(1,3,1,4,3),
                                        Dam = c(NA,2,2,5,6),
                                        WWG = c(4.5,2.9,3.9,3.5,5.0),
                                        PWG = c(6.8,5.0,6.8,6.0,7.5))
```

The table below contains data for pre-weaning gain (WWG) and post-weaning gain (PWG) for `r n_nr_observation` beef calves.

```{r sol12p01table, echo=FALSE, results='asis'}
knitr::kable(tbl_data_sol12p01,
             booktabs = TRUE,
             longtable = TRUE)
```

The genetic variance-covariance matrix $G_0$ between the traits is 

```{r}
mat_g0 <- matrix(data = c(20,18,18,40), nrow = n_nr_trait, byrow = TRUE)
cat("$$\n")
cat("G_0 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_g0, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The residual variance-covariance matrix $R_0$ between the traits is

```{r}
mat_r0 <- matrix(data = c(40,11,11,30), nrow = n_nr_trait, byrow = TRUE)
cat("$$\n")
cat("R_0 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_r0, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


### Your Task
Set up the mixed model equations for a multivariate BLUP analysis and compute the estimates for the fixed effects and the predictions for the breeding values.


### Solution
The matrices $X_1$ and $X_2$ relate records of PWG and WWG to sex effects. For both traits, we have an effect for the male and female sex. Hence the vector $\beta$ of fixed effects corresponds to 

```{r}
vec_beta <- c("\\beta_{M,WWG}", "\\beta_{F,WWG}", "\\beta_{M,PWG}", "\\beta_{F,PWG}")
cat("$$\n")
cat("\\beta = \\left[")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_beta), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


The matrices $X_1$ and $X_2$ are the same and correspond to 

```{r}
mat_x1 <- mat_x2 <- matrix(data = c(1, 0, 
                                    0, 1,
                                    0, 1,
                                    1, 0,
                                    1, 0), nrow = n_nr_observation, byrow = TRUE)
cat("$$\n")
cat("X_1 = X_2 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_x1, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

Combining them to the multivariate version leads to 

$$X = \left[
  \begin{array}{lr}
  X_1  &  0 \\
  0    &  X_2
  \end{array}
\right]$$

```{r}
mat_zero <- matrix(0, nrow = nrow(mat_x1), ncol = ncol(mat_x1))
mat_x <- rbind(cbind(mat_x1, mat_zero), cbind(mat_zero, mat_x2))
cat("$$\n")
cat("X = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_x, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

Using the matrix $X$ together with matrix $R = I_n \otimes R_0$ to get

```{r}
mat_r <- mat_r0 %x% diag(1, n_nr_observation)  
mat_rinv <- solve(mat_r)
mat_xtrinvx <- t(mat_x) %*% mat_rinv %*% mat_x
cat("$$\n")
cat("X^TR^{-1}X = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_xtrinvx, pnDigits = 3), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

Similarly to the fixed effects, we can put together the vector of breeding values $a$. 

```{r}
vec_a1 <- c("a_{1,WWG}",
            "a_{2,WWG}",
            "a_{3,WWG}",
            "a_{4,WWG}",
            "a_{5,WWG}",
            "a_{6,WWG}",
            "a_{7,WWG}",
            "a_{8,WWG}")
vec_a2 <- c("a_{1,PWG}",
            "a_{2,PWG}",
            "a_{3,PWG}",
            "a_{4,PWG}",
            "a_{5,PWG}",
            "a_{6,PWG}",
            "a_{7,PWG}",
            "a_{8,PWG}")
vec_a <- c(vec_a1, vec_a2)
cat("$$\n")
cat("a = \\left[")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_a), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


The design matrices $Z_1$ and $Z_2$ are equal and they link observations to breeding values. 

```{r}
mat_z1zero <- matrix(0, nrow = n_nr_observation, ncol = n_nr_founder)
mat_z1 <- mat_z2 <- cbind(mat_z1zero, diag(1, n_nr_observation))
cat("$$\n")
cat("Z_1 = Z_2 = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_z1, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

$$Z = \left[
  \begin{array}{lr}
  Z_1  &  0 \\
  0    &  Z_2
  \end{array}
\right]$$

```{r}
mat_zzero <- matrix(0, nrow = nrow(mat_z1), ncol(mat_z2))
mat_z <- rbind(cbind(mat_z1, mat_zzero), cbind(mat_zzero, mat_z2))
cat("$$\n")
cat("Z = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_z, pnDigits = 0), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

Together with the numerator relationship matrix $A$ we can get $G = G_0 \otimes A$ and from this $G^{-1} = G_0^{-1} \otimes A^{-1}$

```{r}
ped_sol12p01 <- pedigreemm::pedigree(sire = c(rep(NA, n_nr_founder), tbl_data_sol12p01$Sire), 
                                     dam  = c(rep(NA, n_nr_founder), tbl_data_sol12p01$Dam),
                                     label = as.character(1:n_nr_animal))
mat_ainv <- as.matrix(pedigreemm::getAInv(ped = ped_sol12p01))
cat("$$\n")
cat("A^{-1} = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_ainv, pnDigits = 3), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

```{r}
mat_ginv <- solve(mat_g0) %x% mat_ainv
cat("\\tiny \n")
cat("$$\n")
cat("G^{-1} = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_ginv, pnDigits = 2), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
cat("\\normalsize \n")
```

Using the matrics $X$, $Z$, $R^{-1}$ and $G^{-1}$, we can compute $Z^TR^{-1}X$ and $Z^TR^{-1}Z + G^{-1}$. These matrices define the right-hand side of the mixed model equations. But they are too be to be shown here.

The vector $y$ of observations contains all observations of both traits

```{r}
vec_y1 <- tbl_data_sol12p01$WWG
vec_y2 <- tbl_data_sol12p01$PWG
vec_y <- c(vec_y1, vec_y2)
cat("$$\n")
cat("y = \\left[")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_y), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The right-hand side is computed as

$$
\left[
  \begin{array}{c}
  X^TR^{-1}y \\
  Z^TR^{-1}y
  \end{array}
\right]
$$

The solutions are 

```{r}
### # coefficient matrix
mat_xtrinvz <- t(mat_x) %*% mat_rinv %*% mat_z
mat_ztrinvzginv <- t(mat_z) %*% mat_rinv %*% mat_z + mat_ginv
mat_coeff <- rbind(cbind(mat_xtrinvx, mat_xtrinvz), cbind(t(mat_xtrinvz), mat_ztrinvzginv))
### # right-hand side
mat_rhs <- rbind(t(mat_x) %*% mat_rinv %*% vec_y,
                 t(mat_z) %*% mat_rinv %*% vec_y)
vec_sol <- solve(mat_coeff, mat_rhs)

vec_beta_hat <- c("\\widehat{\\beta_{M,WWG}}", 
                  "\\widehat{\\beta_{F,WWG}}", 
                  "\\widehat{\\beta_{M,PWG}}", 
                  "\\widehat{\\beta_{F,PWG}}")
vec_a1_hat <- c("\\widehat{a_{1,WWG}}",
            "\\widehat{a_{2,WWG}}",
            "\\widehat{a_{3,WWG}}",
            "\\widehat{a_{4,WWG}}",
            "\\widehat{a_{5,WWG}}",
            "\\widehat{a_{6,WWG}}",
            "\\widehat{a_{7,WWG}}",
            "\\widehat{a_{8,WWG}}")
vec_a2_hat <- c("\\widehat{a_{1,PWG}}",
            "\\widehat{a_{2,PWG}}",
            "\\widehat{a_{3,PWG}}",
            "\\widehat{a_{4,PWG}}",
            "\\widehat{a_{5,PWG}}",
            "\\widehat{a_{6,PWG}}",
            "\\widehat{a_{7,PWG}}",
            "\\widehat{a_{8,PWG}}")
vec_a_hat <- c(vec_a1_hat, vec_a2_hat)

vec_hat_unknown <- c(vec_beta_hat, vec_a_hat)
cat("$$\n")
cat("\\left[")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_hat_unknown), collapse = "\n"), "\n")
cat("\\right]")
cat(" = ")
cat("\\left[")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_sol, pnDigits = 4), collapse = "\n"), "\n")
cat("\\right]")
cat("$$\n")
```



\vspace{3ex}

```{r, echo=FALSE, results='asis'}
cat(cnt$out(ps_suffix = "Variance Components Estimation"), "\n")
```
The simplest forM of variance components estimation is based on the residuals of a fitted linear model and is shown in the summary results of the R-function `lm()`. Let us assume that we are given the dataset in the table shown below to which we fit a simple sire model. 


```{r datasol12p02}
tbl_data_sol12p02 <- tibble::data_frame( Animal = c(4, 5, 6, 7, 8),
                                         Sire   = c(2, 1, 1, 2, 1),
                                         WWG    = c(4.5,2.9,3.9,3.5,5.0) )

knitr::kable(tbl_data_sol12p02,
             booktabs  = TRUE,
             longtable = TRUE,
             caption   = "Example Dataset for Variance Components Estimation Based on Residuals Using a Sire Model")

```

The sire model is simplified to have a common mean $\mu$.  For a moment we are setting the sire effects to be fixed effects. This leads to the following model with $var(e) = I * \sigma_e^2$

$$y = X\mu + Z_ss + e$$

Using the above shown dataset we can use the R-function `lm()` to fit this simple linear model. Because, we want to have the sires as fixed effects, we have to convert them into factors before calling `lm()`.

```{r, echo=TRUE, results='markup'}
tbl_data_sol12p02$Sire <- as.factor(tbl_data_sol12p02$Sire)
lm_data_sol12p02 <- lm( WWG ~ 1 + Sire, data = tbl_data_sol12p02 )
summary(lm_data_sol12p02)
```

```{r, echo=FALSE, results='hide'}
res_std_sol12p02 <- sigma(lm_data_sol12p02)
```

From the output of `summary()` we are given the residual standard error to be `r round(res_std_sol12p02, digits = 4)`. This residual standard error is an estimate of $\sigma_e$. The question is where does it come from. The least-squares procedure does not yield this estimate for $\sigma_e$. The answer is that this estimate comes from the residuals $r$ of the model. For our model the vector $r$ of residuals is defined as

$$r = y - X\widehat{\mu} - Z_s\widehat{s}$$

where $\widehat{\mu}$ and $\widehat{s}$ can be taken from the ouput of the `summary()` function. They correspond to

```{r, echo=FALSE, results='asis'}
vec_coeff <- coefficients(lm_data_sol12p02)
n_mu_hat <- vec_coeff[[1]]
vec_sire_hat <- c(0, vec_coeff[[2]])

cat("$$\\widehat{\\mu} = ", n_mu_hat, "$$\n")
cat("$$\n")
cat("\\widehat{s} = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_sire_hat, pnDigits = 4), collapse = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

The estimate $\widehat{\sigma_e^2}$ for $\sigma_e^2$ is obtained by

$$\widehat{\sigma_e^2} = \frac{1}{n-p} \sum_{i=1}^n r_i^2$$

where $n$ is the total number of observations and $p$ is the number of parameters that are estimated by `lm` which is `r length(vec_coeff)` for our sire model. The term $n-p$ is also called __degrees of freedom__ (df). What is given as residual standard error by the output of `summary()` is the square root of $\widehat{\sigma_e^2}$.

### Your Task
Verify for the above given dataset and the proposed sire model the residual standard error given by `summary()` by using the computation based on the residuals shown above.


### Solution
The vector $r$ of residuals can be obtained using the function `residuals()`

```{r, echo=TRUE, results='markup'}
(vec_res <- residuals(lm_data_sol12p02))
```

The degrees of freedom for the residuals ($n-p$) are obtained by the function `df.residual()`

```{r, echo=TRUE, results='markup'}
(n_df_e <- df.residual(lm_data_sol12p02))
```

From this the residual standard error is computed as

```{r, echo=TRUE, results='markup'}
(n_res_sd <- sqrt(sum(vec_res^2) / n_df_e))
```

The same result can be obtained using the function `sigma()`

```{r, echo=TRUE, results='markup'}
sigma(lm_data_sol12p02)
```






