---
title: "Inverse Numerator Relationship Matrix and Inbreeding"
author: "Peter von Rohr"
date: "16 November 2018"
output:
  beamer_presentation:
header-includes:
 \usepackage{longtable}
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "asis")
knitr::knit_hooks$set(hook_convert_odg = rmddochelper::hook_convert_odg)
```

## Inverse of $A$ Based on $L$ and $D$

* Matrix $A$ was decomposed into $A = L \cdot D \cdot L^T$
* Get $A^{-1}$ as $A^{-1} = (L^T)^{-1} D^{-1} L^{-1}$
* $D^{-1}$ is diagonal again with elements

$$(D^{-1})_{ii} = 1/(D)_{ii}$$

## Inverse of $L$

* Compute $m$ based on the two decompositions of $a$

$$a = P \cdot a + m \quad \text{and} \quad a = L \cdot m$$

* Solve both for $m$ and set them equal

$$m = a - P \cdot a = (I-P)\cdot a \quad \text{and} \quad m = L^{-1} \cdot a$$

$$(I-P) \cdot a = L^{-1} \cdot a$$
and

$$L^{-1} = I-P$$

## Example

```{r showsimplepedigree, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
### # first six animals from Goetz p. 83
suppressPackageStartupMessages( library(pedigreemm) )
n_nr_ani_ped <- 5
n_nr_parent <- 3
tbl_ped_simple <- dplyr::data_frame(Calf = c(1:n_nr_ani_ped),
                             Sire = c(NA, NA, NA, 1, 3),
                             Dam  = c(NA, NA, NA, 2, 2))
### # pedigreemm
ped_simple <- pedigree(sire = tbl_ped_simple$Sire, dam = tbl_ped_simple$Dam, label = as.character(1:n_nr_ani_ped))
matA_simple <- as.matrix(getA(ped = ped_simple))
matAinv_simple <- as.matrix(getAInv(ped = ped_simple))
### # LDL decomposition based on cholesky
matR <- t(chol(matA_simple))
### # matS = sqrt(matD)
matD <- diag(Dmat(ped = ped_simple), n_nr_ani_ped)
matS <- sqrt(matD)
matL <- matR %*% solve(matS)
### # inverse of D
matDinv <- solve(matD)
### # matrix P
matP <- diag(1, n_nr_ani_ped) - solve(matL)
matLinv <- diag(1, n_nr_ani_ped) - matP
### # decomposition of AInv
matLinvtmatDinv <- crossprod(matLinv, matDinv)
```

```{r tabpedsimpleexample, echo=FALSE}
knitr::kable(tbl_ped_simple,
             booktabs = TRUE,
             longtable = TRUE)
```


## Matrix $D^{-1}$

* Because $D$ is diagonal

```{r examplematlmatd, echo=FALSE, results='asis'}
cat("$$\n")
cat("D = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matD, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

* We get $D^{-1}$ as
```{r examplematlmatdinv, echo=FALSE, results='asis'}
cat("$$\n")
cat("D^{-1} = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matDinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


## Matrix $L^{-1}$

* Use $L^{-1} = I-P$
* Matrix $P$ from simple decomposition

```{r examplematp}
cat("$$\n")
cat("P = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matP, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```

* Therefore

```{r}
cat("$$\n")
cat("L^{-1} = I-P = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matLinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
```


## Decomposition of $A^{-1}$ I

```{r}
cat("$$\n")
cat("A^{-1} = (L^{-1})^T \\cdot D^{-1} \\cdot L^{-1}")
cat("$$\n")

cat("$$\n")
cat("(L^{-1})^T \\cdot D^{-1}")
cat("$$\n")

cat("\\small \n $$\n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = t(matLinv), pnDigits = 1), sep = "\n"), "\n")
cat("\\right] \\cdot \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matDinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
cat("$$ = \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matLinvtmatDinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$ \\normalsize \n")
```


## Decomposition of $A^{-1}$ II

```{r}
cat("$$\n")
cat("A^{-1} = (L^{-1})^T \\cdot D^{-1} \\cdot L^{-1}")
cat("$$\n")

cat("\\small \n $$\n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matLinvtmatDinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right] \\cdot \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matLinv, pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$\n")
cat("$$ = \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = crossprod(t(matLinvtmatDinv),matLinv), pnDigits = 1), sep = "\n"), "\n")
cat("\\right]\n")
cat("$$ \\normalsize \n")
```


## Henderson's Rules

* Both Parents Known
    + add $2$ to the diagonal-element $(i,i)$
    + add $-1$ to off-diagonal elements $(s,i)$, $(i,s)$, $(d,i)$ and $(i,d)$
    + add ${1\over 2}$ to elements $(s,s)$, $(d,d)$, $(s,d)$, $(d,s)$

* Only One Parent Known
    + add ${4\over 3}$ to diagonal-element $(i,i)$
    + add $-{2\over 3}$ to off-diagonal elements $(s,i)$, $(i,s)$
    + add ${1\over 3}$ to element $(s,s)$

* Both Parents Unknown
    + add $1$ to diagonal-element $(i,i)$

* Valid without inbreeding 


## Inbreeding

* Elements in matrix $D$ depend on coefficients of inbreeding
* Recap: From the simple decomposition of $a$, we derived 

\begin{align}
var(m_i)  &=  \left({1\over 2} - {1\over 4}(F_s + F_d)\right) \sigma_a^2  \notag \\
          &=  \left({1\over 2} - {1\over 4}(A_{ss} - 1 + A_{dd} - 1)\right) \sigma_a^2  \notag \\
          &=  \left(1 - {1\over 4}(A_{ss} + A_{dd}) \right) \sigma_a^2  \notag \\
          &=  (D)_{ii} \sigma_a^2 \notag
\end{align}

$$\rightarrow  (D)_{ii}  = \left({1\over 2} - {1\over 4}(F_s + F_d)\right) = \left(1 - {1\over 4}(A_{ss} + A_{dd}) \right)$$ 


## Computation of Coefficients of Inbreeding

* Observation: Coefficients of inbreeding $F_s$ and $F_d$ can be read from $(A)_{ss}$ and $(A)_{dd}$ of $A$
* Cannot setup $A$ to just get inbreeding coefficients
* More efficient method required
* __Cholesky__ decomposition of $A$

$$A = R\cdot R^T$$

where $R$ is a lower triangular matrix

__Hint__: Function `chol(A)` in R gives matrix $R^T$


## Cholesky Decomposition

* Diagonal elements $(A)_{ii}$ of $A$ are the sum of the squared elements of one row of $R$

\begin{equation}
(A)_{ii} = \sum_{j=1}^i (R)_{ij}^2 \notag
\end{equation}

* Example

```{r SmallExCholA, eval=TRUE, echo=FALSE, results='asis'}
nAnzAni <- 3
matA <- rmddochelper::matGetMatElem(psBaseElement = "(A)", pnNrRow = nAnzAni, pnNrCol = nAnzAni)
matR <- rmddochelper::matLowerTri(psBaseElement = "(R)", pnNrRow = nAnzAni, pnNrCol = nAnzAni)
cat("\\tiny \n $$\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matA), collapse = "\n"))
cat("\\right] \n")
cat(" = \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matR), collapse = "\n"))
cat("\\right] \n")
cat(" \\cdot \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = t(matR)), collapse = "\n"))
cat("\\right] \n")
cat("$$\n \\normalsize \\")
```


## Recursive Computation of $R$

* Let us write the matrix $R$ as a product of two matrices $L$ and $S$:

$$R = L \cdot S$$
where $L$ is the same matrix as in the `LDL`-decompositon and $S$ is a diagonal matrix.

* Compute $A$ as

$$A = R \cdot R^T = L \cdot S \cdot S \cdot L^T = L \cdot D \cdot L^T$$

* Hence

$$D = S \cdot S \quad \rightarrow \quad (S)_{ii} = \sqrt{(D)_{ii}}$$

## Example

```{r RlsDecompEx, eval=TRUE, echo=FALSE, results='asis'}
matL <- rmddochelper::matLowerTri(psBaseElement = "(L)", pnNrRow = nAnzAni, pnNrCol = nAnzAni, pvecDiag = 1)
matS <- rmddochelper::matDiag(psBaseElement = "(S)", pnNrRow = nAnzAni, pnNrCol = nAnzAni)
cat("\\tiny \n $$ \n \\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matR), collapse = "\n"))
cat("\\right]  = \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matL), collapse = "\n"))
cat("\\right]  \\cdot \n")
cat("\\left[")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = matS), collapse = "\n"))
cat("\\right]\n $$ \n \\normalsize \\")
```


* Diagnoal elements $(R)_{ii} = (S)_{ii}$

* Because $(S)_{ii} = \sqrt{(D)_{ii}}$, diagonal elements $(R)_{ii}$ of matrix $R$ can be computed as 

$$(R)_{ii} =  (S)_{ii} = \sqrt{(D)_{ii}} = \sqrt{\left(1 - {1\over 4}(A_{ss} + A_{dd}) \right)}$$

* $A_{ss}$ and $A_{dd}$ are
    + $0$ if $s$ and $d$ are unknown (NA) or
    + have been computed before


## Offdiagonal Elements of $R$

* Offdiagnoal elements $(R)_{ij}$ of $R$ are computed as 

$$(R)_{ij} = (L)_{ij} * (S)_{jj}$$

* Use property of $L$: $L_{ij} = {1\over 2}((L)_{sj} + (L)_{dj})$ if $s$ and $d$ are parents of $i$

\begin{align}
(R)_{ij}  &=  (L)_{ij} * (S)_{jj} \notag \\
          &=  {1\over 2} \left[ (L)_{sj} + (L)_{dj} \right] * (S)_{jj} \notag \\
          &=  {1\over 2} \left[ (L)_{sj} * (S)_{jj} + (L)_{dj} * (S)_{jj} \right]  \notag \\
          &=  {1\over 2} \left[ (R)_{sj} + (R)_{dj} \right] \notag
\end{align}


## Example Pedigree

```{r showextpedigree, echo=FALSE, results='asis'}
n_nr_ani_ped <- 6
n_nr_parent <- 3
tbl_ped_ext <- dplyr::data_frame(Calf = c(1:n_nr_ani_ped),
                             Sire = c(NA, NA, NA, 1, 3, 4),
                             Dam  = c(NA, NA, NA, 2, 2, 5))
### # pedigreemm
ped_ext <- pedigree(sire = tbl_ped_ext$Sire, dam = tbl_ped_ext$Dam, label = as.character(1:n_nr_ani_ped))
matA_ext <- as.matrix(getA(ped = ped_ext))
matAinv_ext <- as.matrix(getAInv(ped = ped_ext))
### # LDL decomposition based on cholesky
matR_ext <- t(chol(matA_ext))
### # matS = sqrt(matD)
matD_ext <- diag(Dmat(ped = ped_ext), n_nr_ani_ped)
matS_ext <- sqrt(matD_ext)
matL_ext <- matR_ext %*% solve(matS_ext)
### # show table
knitr::kable(tbl_ped_ext, 
             booktabs  = TRUE,
             longtab   = TRUE)
```


## Computations

* Compute diagonal elements $(A)_{ii}$ of $A$ to get $F_i$
* Prerequisite: Pedigree sorted such that parents before progeny
* Start with $(A)_{11}$

\begin{align}
(A)_{11}  &= (R)_{11}^2 = (D)_{11} = \left(1 - {1\over 4}(A_{NANA} + A_{NANA}) \right) = 1 \notag 
\end{align}

* $(A)_{22}  = (R)_{21}^2 + (R)_{22}^2 = 0 + 1 = 1$

* $(A)_{33}  = (R)_{31}^2 + (R)_{32}^2 + (R)_{33}^2 = 0 + 0 + 1 = 1$


## Animals With Known Parents

\begin{align}
(A)_{44}  &=  (R)_{41}^2 + (R)_{42}^2 + (R)_{43}^2 + (R)_{44}^2 \notag \\
          &=  ({1\over 2}(R_{11} + R_{21}))^2 + ({1\over 2}(R_{12} + R_{22}))^2 + ({1\over 2}(R_{13} + R_{23}))^2 \notag \\
          & + \left(1 - {1\over 4}(A_{11} + A_{11}) \right) \notag \\
          &=  {1\over 4} + {1\over 4} + {1\over 2} = 1\notag
\end{align}

* $(A)_{55}$

* $(A)_{66}$
