---
title: "Sending E-Mail Announcements To Students"
author: "Peter von Rohr"
date: "2018-09-18"
output: html_notebook
---

## Disclaimer
This notebook contains a protocol on all the steps that were taken to send out e-mail announcements to students. The process of sending out e-mails with `gmailr` is taken from the following two resources

1. https://github.com/jennybc/send-email-with-r and
2. https://github.com/jimhester/gmailr


## Prerequisites
On the google developers console (console.developers.google.com), a new project was created and the GMail API was enabled. As the next step, credentials were created and the resulting json-file that contains the credentials was downloaded. More details on how to do that can be taken from https://github.com/jimhester/gmailr.


## Getting Started
We are following the tutorial on https://github.com/jennybc/send-email-with-r to create and send the announcement e-mails.

### Required packages
The following packages are used

```{r}
vec_pkgs <- c("gmailr", "readr", "dplyr", "purrr")
vec_inst_pkgs <- installed.packages()
for (p in vec_pkgs){
  if (!p %in% vec_inst_pkgs)
    install.packages(pkgs = p, repos = "https://cran.rstudio.com")
}
```


 
### Dry Run
We start with a test

```{r}
### # name of the secret file
s_secret_file <- "lbgfs2018.googleusercontent.com.json"
# that secret file must exist, o/w we stop here
stopifnot(file.exists(s_secret_file))
```

Once the secret file is set, we can start

```{r}
### # start the test
suppressPackageStartupMessages(library(gmailr))
use_secret_file(s_secret_file)
```

The following code snippet sends a simple message to a defined address. This test is only run once, hence, we are not evaluating the following code in subsequent runs.

```{r, eval=FALSE}
test_email <- mime(
    To = "peter.vonrohr19680917@bluewin.ch",
    From = "peter.vonrohr@gmail.com",
    Subject = "this is just a gmailr test",
    body = "Can you hear me now?")
send_message(test_email)
```

The first time when running this, we are presented with the following question:

```
Use a local file ('.httr-oauth'), to cache OAuth access credentials between R sessions?

1: Yes
2: No

```

Entering 1 as the answer produces the following output and opens a browser where the project must be associated with the given e-mail address. As soon as this process is done, the browser window can be closed again.

```
Adding .httr-oauth to .gitignore
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
Id: 165ed3158db06220
To: 
From: 
Date: 
Subject: 
```
Checking the inbox of the address specified in the above "To"-field must result in a new message. If that message is not received, then something is wrong.


## Compose and Send E-Mails
From the teaching system all student names were exported and just the columns with the names and the e-mails were retained. Then this information is saved in a .csv-file. This file is read into a dataframe.

```{r}
s_adr_file <- "addresses_lbgfs2018.csv"
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(readr))

addresses <- read_csv(s_adr_file)
```

From the E-Mail addresses, we create the logins

```{r}
addresses <- addresses %>% rowwise() %>% mutate(login = strsplit(`E-Mail`, split = "@")[[1]][1])
```

We are adding a test student with artificial data, but with a valid E-Mail address.

```{r}
addresses <- addresses %>%
  add_row(Familienname="Student", Vorname="Test", Rufname="Test", Nummer=NA, Sprache=NA, `E-Mail`="peter.vonrohr19680917@bluewin.ch", login="pvrlbgfs2018")
#addresses
```

Write all logins to a separate file

```{r}
cat(addresses$login, file = 'login.txt', sep = "\n")
```

The file `login.txt` with the usernames for the students is uploaded to the server. We are using the script `create_user.sh` from bashtools to create a new user. For testing purposes, the user `pvrlbgfs2018` is created. It is important to create users with usernames that did not exist previously, because deleteing and re-creating old usernames cause problems when trying to login to the RStudio-server via the browser.

The following bash-command is used to create all users

```
cat login.txt | while read e;do echo $e;sudo /opt/bashtools/sysadmin/create_user.sh -u $e -s /bin/bash;sleep 2;done
```

Usernames and passwords are collected into a single file called `access.txt` with the following command.

```
ls -1 .*.pwd | while read f;do echo $f;cat $f >> access.txt;sleep 2;done
```

This file is downloaded and used to send out E-Mails with passwords to each student. To do that we have to first read the access data.

```{r}
access <- read_csv(file = "access.txt", col_names = c("login","password"))
#access
```

The two tibbles `addresses` and `access` are joined 

```{r}
my_dat <- left_join(addresses, access)
# <- my_dat %>% filter(login == "pvrlbgfs2018")
#my_dat
```

A new tibble is created which contains the relevant fields of the E-Mails.

```{r}
email_sender <- 'Peter von Rohr<peter.vonrohr@gmail.com>' # your Gmail address
body <- "Dear %s %s,

Your access for the exercise platform is: %s

Good luck and enjoy using the platform!

Best regards, Peter von Rohr
"
edat <- my_dat %>%
    mutate(
        To = `E-Mail`,
        From = email_sender,
        Subject = 'Access to exercise platform',
        body = sprintf(body, Vorname, Familienname, password)) %>%
    select(To, From, Subject, body)
write_csv(edat, "composed-emails.csv")
```

Now `gmailr::mime()` is used to construct the E-Mails.

```{r}
emails <- edat %>%
  pmap(mime)
str(emails, max.level = 2, list.len = 2)
```

## Sending the E-Mails

```{r}
safe_send_message <- safely(send_message)
sent_mail <- emails %>% 
  map(safe_send_message)

saveRDS(sent_mail, "lbgfs2018_sent-emails.rds")
```

Checking for errors

```{r}
errors <- sent_mail %>% 
  transpose() %>% 
  .$error %>% 
  map_lgl(Negate(is.null))
sent_mail[errors]
```

The general E-Mail sent to the students, looks as follows

```
Dear Students,
my name is Peter von Rohr. You are receiving this E-Mail, because you registered to the course Livestock Breeding and Genomics. 

The website for this course is available at: https://charlotte-ngs.github.io/LBGFS2018/

Exercises for this course can be done on a separate platform (http://r4tea.rteastem.org:8787) which consists of an RStudio server. The username to access the 
platform is the part of your E-Mail address before the `@`-sign. The password is sent to you in a separate E-Mail. 

I am looking forward to seeing you on Friday at 9:15.
Best regards and have a nice day,
Peter von Rohr

```