---
title: "Sending E-Mail Announcements To Students"
author: "Peter von Rohr"
date: "2018-09-18"
output: html_notebook
---

## Disclaimer
This notebook contains a protocol on all the steps that were taken to send out e-mail announcements to students. The process of sending out e-mails with `gmailr` is taken from the following two resources

1. https://github.com/jennybc/send-email-with-r and
2. https://github.com/jimhester/gmailr


## Prerequisites
On the google developers console (console.developers.google.com), a new project was created and the GMail API was enabled. As the next step, credentials were created and the resulting json-file that contains the credentials was downloaded. More details on how to do that can be taken from https://github.com/jimhester/gmailr.


## Getting Started
We are following the tutorial on https://github.com/jennybc/send-email-with-r to create and send the announcement e-mails.

### Required packages
The following packages are used

```{r}
vec_pkgs <- c("gmailr", "readr", "dplyr", "purrr")
vec_inst_pkgs <- installed.packages()
for (p in vec_pkgs){
  if (!p %in% vec_inst_pkgs)
    install.packages(pkgs = p, repos = "https://cran.rstudio.com")
}
```


 
### Dry Run
We start with a test

```{r}
### # name of the secret file
s_secret_file <- "lbgfs2018.googleusercontent.com.json"
# that secret file must exist, o/w we stop here
stopifnot(file.exists(s_secret_file))
```

Once the secret file is set, we can start

```{r}
### # start the test
suppressPackageStartupMessages(library(gmailr))
use_secret_file(s_secret_file)
```

The following code snippet sends a simple message to a defined address. This test is only run once, hence, we are not evaluating the following code in subsequent runs.

```{r, eval=FALSE}
test_email <- mime(
    To = "peter.vonrohr19680917@bluewin.ch",
    From = "peter.vonrohr@gmail.com",
    Subject = "this is just a gmailr test",
    body = "Can you hear me now?")
send_message(test_email)
```

The first time when running this, we are presented with the following question:

```
Use a local file ('.httr-oauth'), to cache OAuth access credentials between R sessions?

1: Yes
2: No

```

Entering 1 as the answer produces the following output and opens a browser where the project must be associated with the given e-mail address. As soon as this process is done, the browser window can be closed again.

```
Adding .httr-oauth to .gitignore
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
Id: 165ed3158db06220
To: 
From: 
Date: 
Subject: 
```
Checking the inbox of the address specified in the above "To"-field must result in a new message. If that message is not received, then something is wrong.


## Compose and Send E-Mails
From the teaching system all student names were exported and just the columns with the names and the e-mails were retained. Then this information is saved in a .csv-file. This file is read into a dataframe.

```{r}
s_adr_file <- "addresses_lbgfs2018.csv"
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(readr))

(addresses <- read_csv(s_adr_file))
```

From the E-Mail addresses, we create the logins

```{r}
addresses <- addresses %>% rowwise() %>% mutate(login = strsplit(`E-Mail`, split = "@")[[1]][1])
addresses
```


