---
title: "Generate Data for Problem 2 in Exercise 8"
output:
  html_document:
    df_print: paged
date: '2018-11-11'
---

## Disclaimer
This notebook shows how data for problem 2 in exercise 8 are generated.


## The General Setup
The data generated depends on

* a general mean $\mu$ or an intercept $\beta_0$
* a herd effect with two levels
* a random animal effect with a given variance-covariance structure specified by the pedigree
* a random residual effect with a given variance component.

```{r nrobsdef, echo=FALSE, results='hide'}
n_nr_obs <- 6
```

The number of observations are fixed to `r n_nr_obs`. 

## The Fixed Effects
For the fixed effect, we are specifying the matrix $X$ and the vector $\beta$. 

```{r betadef, echo=FALSE, results='asis'}
vec_beta <- c(10.34, 5.2, -3.92)
cat("$$\n")
cat("\\beta = \\left[\n")
cat(paste(rmddochelper::sConvertVectorToLaTexArray(pvec_avector = vec_beta), collapse = "\n"),"\n")
cat("\\right]\n")
cat("$$\n")
```

The incidence matrix $X$ is defined as

```{r xdef, echo=FALSE, results='asis'}
mat_x <- matrix(c(1, 1, 0,
                  1, 1, 0,
                  1, 1, 0,
                  1, 0, 1,
                  1, 0, 1,
                  1, 0, 1), nrow = n_nr_obs, byrow = TRUE)
cat("$$\n")
cat("X = \\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_x, pnDigits = 0), collapse = "\n"),"\n")
cat("\\right]\n")
cat("$$\n")
```



## The Variances and the Random Effects
There are two random effects

1. residuals $e$
2. additive genetic breeding value $a$


### Residuals
```{r sigmaedef, echo=FALSE, results='hide'}
n_sigmae <- 24
```

The random residuals are given by a fixed variance component $\sigma_e^2 = `r n_sigmae`$. The residual compontents are generated from a normal distribution with mean $0$ and variance equal to $\sigma_e^2$. 


### Breeding Values
```{r sigmaadef, echo=FALSE, results='hide'}
n_sigmaa <- 8
```

The genetic additive variance is set to $\sigma_a^2 = `r n_sigmaa`$. The variance-covariance structure is given by the pedigree which is shown in the following table.

```{r peddef, echo=FALSE, results='asis'}
tbl_ped_ex8p1 <- tibble::data_frame(Animal = c(5, 6, 7, 8, 9, 10),
                                    Sire   = c(1, 1, 4, 4, 4, 4),
                                    Dam    = c(2, 3, 5, 5, 6, 6))

n_nr_ani_ex8p1pedext <- max(tbl_ped_ex8p1$Animal)
n_nr_founder_ex8p1pedext <- min(tbl_ped_ex8p1$Animal) - 1
tbl_ped_exp8p1ext <- tibble::data_frame(Animal = c(1:n_nr_ani_ex8p1pedext),
                                        Sire   = c(rep(NA, n_nr_founder_ex8p1pedext), tbl_ped_ex8p1$Sire),
                                        Dam    = c(rep(NA, n_nr_founder_ex8p1pedext), tbl_ped_ex8p1$Dam))
knitr::kable(tbl_ped_exp8p1ext,
             booktabs  = TRUE,
             longtable = TRUE,
             caption   = "Extended Pedigree")
```

For the founders, the breeding values are generated from a normal distribution with mean $0$ and variance equal to $\sigma_a^2$. For animals $i$ with parents, the breeding values $a_i$ are computed by

$$a_i = {1\over 2}a_s + {1\over 2}a_d + m_i$$

where $a_s$ and $a_d$ are breeding values of parents $s$ and $d$ and $m_i$ is a random compnent with variance $var(m_i)$ which corresponds to the product of $\sigma_a^2$ times the $i$-th diagonal element of matrix $D$ from the `LDL`-decomposition of $A$. The matrix $D$ can be obtained using the function `pedigreemm::Dmat()`.

```{r getafromped, echo=FALSE, results='asis'}
suppressPackageStartupMessages( library(pedigreemm) )
ped <- pedigree(sire = tbl_ped_exp8p1ext$Sire, dam = tbl_ped_exp8p1ext$Dam, label = as.character(1:n_nr_ani_ex8p1pedext))
mat_d <- as.matrix(pedigreemm::Dmat(ped = ped))
cat("$$\n")
cat("diag(D) = \\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_d, pnDigits = 2), collapse = "\n"),"\n")
cat("\\right]\n")
cat("$$\n")
```

The design matrix $Z$ will be used to link breeding values to observations. The number of rows of matrix $Z$ corresponds to the number of observations and the number of columns is equal to the total number of animals in the pedigree.

```{r matzdef, echo=FALSE, results='asis'}
mat_z <- cbind(matrix(0, nrow = n_nr_obs, ncol = n_nr_founder_ex8p1pedext), diag(1, nrow = n_nr_obs, ncol = n_nr_obs))
cat("$$\n")
cat("Z = \\left[\n")
cat(paste(rmddochelper::sConvertMatrixToLaTexArray(pmatAMatrix = mat_z, pnDigits = 0), collapse = "\n"),"\n")
cat("\\right]\n")
cat("$$\n")
```


## Generate Observations
Phenotypic observations are generated according to the standard animal model

$$y = X \beta + Za + e$$

Except for the vector of breeding values $a$, all other components are easy to generate using built-in R-functions or can readily be computed using matrix-vector multiplication. The components of vector $a$ have this internal structure which depends on the pedigree. Hence we try to come up with a function that generates all components of vector $a$.


```{r}
generate_bv <- function(ptbl_ped, pn_sigmaa){
  ### # get matrix D based on pedigree
  ped <- pedigreemm::pedigree(sire = ptbl_ped$Sire, dam = ptbl_ped$Dam, label = as.character(ptbl_ped$Animal))
  diag_mat_d <- as.matrix(pedigreemm::Dmat(ped = ped))
  
  ### # loop over animals in pedigree and generate the breeding values
  for (idx in seq_along(ptbl_ped$Animal)){
    ### # check on parents
    sire_bv <- 0
    if (!is.na(ptbl_ped$Sire[idx]))
      sire_bv <- 
  }
  ### # generate breeding values for founders
  
}
```


## Further Experiments
How can we find the founders

