---
title: Livestock Breeding and Genomics - Exercise 5
author: Peter von Rohr
date: 2018-10-26
#output: pdf_document 
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

```{r PrepareBeefExample, results='hide'}
n_nr_sire <- 3
n_nr_dam <- 8
n_nr_parents <- n_nr_sire + n_nr_dam
n_nr_offspring <- 16
n_nr_animals <- n_nr_parents + n_nr_offspring
dam_id <- rep(4:11,2)

tbl_beef_data <- dplyr::data_frame(Animal = (n_nr_parents + 1):n_nr_animals,
                                   Sire = c(rep(1,8), rep(2,6), rep(3,2)),
                                   # Dam = dam_id[order(dam_id)],
                                   # Herd = c(rep(1,4), rep(2,4), rep(1,4), rep(2,4)),
                                   `Weaning Weight`= c(2.61,2.31,2.44,2.41,2.51,2.55,2.14,2.61,2.34,1.99,3.1,2.81,2.14,2.41,2.54,3.16))
# names(tbl_beef_data) <- c("Animal", "Sire", "Weaning Weight")
n_nr_observation <- nrow(tbl_beef_data)

### # parameters
h2 <- .25
n_var_p <- round(var(tbl_beef_data$`Weaning Weight`), digits = 4)
n_var_g <- round(h2 * n_var_p, digits = 4)
n_pop_mean <- round(mean(tbl_beef_data$`Weaning Weight`), digits = 2)
```

## Prediction of Breeding Values Using the Regression Method

We are using the dataset shown in Table \@ref(tab:TabBeefExample) for this exercise. For the animals listed in Table \@ref(tab:TabBeefExample), the weaning weight  (in 100kg) was observed as phenotypic records. The following parameters are associated with the observed data

* The population mean is assumed to be equal to the average of all observations: $\mu = `r n_pop_mean`$
* The phenotypic variance is assumed to correspond to the empirical variance from the observations and corresponds to $\sigma_p^2 = `r n_var_p`$
* The heritability is assumed to be $h^2 = `r h2`$
* The genetic-additive variance can be computed as $\sigma_a^2 = h^2 * \sigma_p^2 = `r h2` * `r n_var_p` = `r n_var_g`$


```{r TabBeefExample}
knitr::kable(tbl_beef_data, 
             booktabs = TRUE, 
             longtable = TRUE,
             caption = "Example Data Set To Predict Breeding Values")
```


## Problem 1: Own performance
Compute the predicted breeding values and the reliabilities for the animals listed in Table \@ref(tab:TabBeefExample). Compare the ranking of the animals according to their phenotypic values and according to their predicted breeding values. Compare the reliabilities of the predicted breeding values.  


### Your Solution



## Problem 2: Predicted Breeding Values Based on Progeny Records
Compute the predicted breeding values and the reliabilities for the sires based on the progeny records. We are assuming that all progeny for a given sire are half-sibbs. Compare the ranking of the sires according to the average progeny performance values and according to the predicted breeding values.


### Your Solution


